\input cwebmac
\def\title{UTFpatgen}
\def\utfpatgen{{\tt utfpatgen}}
\def\patgen{{\tt patgen}}
\def\lefthyphenminpar{{\tt lefthyphenmin}}
\def\righthyphenminpar{{\tt righthyphenmin}}
\def\hyphstartpar{{\tt hyph\_start}}
\def\hyphfinishpar{{\tt hyph\_finish}}
\def\patstartpar{{\tt pat\_start}}
\def\patfinishpar{{\tt pat\_finish}}
\def\goodwtpar{{\tt good\_wt}}
\def\badwtpar{{\tt bad\_wt}}
\def\threshpar{{\tt thresh}}


\N{0}{1}Introduction.
This is \utfpatgen{} -- reimplementation of the classic \patgen{} program for
pattern generation. With \utfpatgen, we
intend to overcome several limitations of the original, such as the number of
hyphenation levels possible, inability to
use some reserved characters in dictionary, and most importantly, we enable
native usage of the UTF-8 encoding in
dictionaries that are no longer limited by the fixed number of lowercase
characters permitted by \patgen.

We provide \utfpatgen{} open-source and free of charge under MIT license.
Please note that there is no warranty
and despite our greatest effort, the program may contain bugs.

\fi

\N{0}{2}Terminology.
Before diving into the implementation part of the program, it is useful to
mention several terms occurring frequently
throughout the text. It may come useful especially to those who are not
thoroughly familiar with \patgen. The
definitions here are mostly informal and intended to ease reader's
understanding of the topic.

{\bf Hyphenation} is a process of splitting words so that they fit better to
the paragraphs when typeset. It follows
the linguistic rules set for the language in which the text is written.

{\bf Pattern} is a sequence of characters containing hyphenation information
tied to some position ({\bf dot
position}) on the edges of the pattern or in between the characters. Whether
this information tells that the position
should or should not be hyphenated, we distinguish between {\bf hyphenating}
and {\bf inhibiting patterns}.

{\bf Hyphen} (in context of \patgen{} and \utfpatgen) is a mark between two
characters holding the information whether
that position should be split during hyphenation and whether this split was
detected by current patterns. There are
therefore 4 types of hyphens: {\bf NO\_HYF} (not in the data, not marked), {\bf
MISS\_HYF} (in the data, not
marked), {\bf BAD\_HYF} (not in the data, marked), and {\bf GOOD\_HYF} (in the
data, marked). Strictly speaking,
there is a hyphen between each pair of neighboring characters in the data, but
NO\_HYF marking is omitted implicitly.

{\bf Supporting occurences} of a pattern in the data are those that are in
favor of the hyphenation information it
holds. For hyphenating patterns, these are the cases where the word has a MISS%
\_HYF on the dot position pointed by the
pattern. For inhibiting patterns, dot position with a BAD\_HYF is a supporting
one. {\bf Contradicting occurences}
are those that would break a correct hyphen, thus NO\_HYF for hyphenating, and
GOOD\_HYF for inhibiting patterns. Note
that a occurence of a pattern can be neither supporting nor contradicting.

{\bf Hyphenation level} marks the strength of a hyphenation. It is represented
by a non-negative integer and creates
a hierarchy of patterns. Pattern with higher level always takes precedence over
any pattern with lower level. At odd
levels, the patterns are hyphenating, at even levels inhibiting.

{\bf Trie} is a data structure for storing n-ary trees. It can implemented
using an associated array with highly
effective insertion, deletion and lookup. Furthermore, we can condense it via {%
\it packing} to save space.

\fi

\N{0}{3}Algorithm overview.
In general, we tried to adhere as tightly as possible to the original ideas of %
\patgen. Therefore, you may find the
names and functions similar to those in the \patgen{} technical report. We have
nevertheless decided to rewrite several
parts of the algorithm in more "modern" way to improve its readability and
testability. The main points to mention are:

\item{$\bullet$} the algorithm is implemented in CWEB (C being the laguage of
the program), not WEB (with Pascal),
\item{$\bullet$} instead of statically defining the sizes of structures (tries,
buffers, etc.), these are allocated and
reallocated dynamically, allowing for greater flexibility and possibly space
savings,
\item{$\bullet$} global variables are localized as much as possible,
\item{$\bullet$} {\tt goto} statements are eliminated.

You may also find a few unit tests appended to the code. These are by no means
exhaustive, but feel free to run them
and add your own.

We decided to present \utfpatgen{} in top-down fashion -- starting with the
full overview and moving to details in later
sections. Same goes for the structures which have their own dedicated sections.

\fi

\N{1}{4}Dependencies.
All external libraries used in \utfpatgen{} come from the standard C package,
so we hope it to be widely portable without
greater trouble. All in all, we use fixed-size types from {\tt <stdint.h>} and
{\tt <stdbool.h>}, IO support from
{\tt <stdio.h>}, string manipulation methods from {\tt <string.h>}, and memory
management provided by
{\tt <stdlib.h>}.

\Y\B\4\X4:Library includes\X${}\E{}$\6
\8\#\&{include} \.{"utfpatgen.h"}\6
\8\#\&{include} \.{<string.h>}\par
\U5.\fi

\N{1}{5}Main method.
The general flow of the program is rather simple: initialize structures, get
the parameters, generate patterns, and
optionally hyphenate the dictionary. For sure, it gets more complicated the
deeper we dive.

\Y\B\X4:Library includes\X\6
\8\#\&{ifndef} \.{TEST}\7
\1\1\&{int} \\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,])\2%
\2{}$\6
${}\{{}$\1\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{argc};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{strcmp}(\\{argv}[\|i],\39\.{"--help"})\E\T{0}){}$\5
${}\{{}$\1\6
\\{print\_help}(\,);\6
\&{return} \.{EXIT\_SUCCESS};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{strcmp}(\\{argv}[\|i],\39\.{"--version"})\E\T{0}){}$\5
${}\{{}$\1\6
\\{print\_version}(\,);\6
\&{return} \.{EXIT\_SUCCESS};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{print\_version}(\,);\6
\X6:Initialization sequence\X;\6
\X7:Level range specification\X;\6
\X8:Pattern generation\X;\6
\X11:Final pass\X;\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_SUCCESS};\6
\4${}\}{}$\2\6
\8\#\&{endif}\par
\fi

\N{1}{6}Initialization sequence.
First, the input parameters provided to the program call are read, validated
and processed. Unless asking for help
({\tt --help}) or version printout ({\tt --version}), \utfpatgen{} takes
exactly 4 inputs, representing 4 files:

\item{$\bullet$} {\bf Dictionary file}: contains set of hyphenated words.
\item{$\bullet$} {\bf Patterns file}: stores patterns generated in previous
runs.
\item{$\bullet$} {\bf Output file}: where the patterns will be stored after the
run.
\item{$\bullet$} {\bf Translate file}: contains the mapping of characters from
the dictionary and dictionary-specific
parameters.

The required formats of these files are the same as for the \patgen{} program
and are discussed in dedicated sections.

\Y\B\4\X6:Initialization sequence\X${}\E{}$\6
\&{struct} \\{params} ${}{*}\\{params}\K\\{init\_params}(\,);{}$\7
\&{if} ${}(\\{params}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{parse\_input}(\\{argv},\39\\{argc},\39\\{params})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\7
\&{struct} \\{translate\_table} ${}{*}\\{tt}\K\\{init\_tr\_table}(\T{256},\39%
\T{128});{}$\7
\&{if} ${}(\\{tt}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{read\_translate}(\\{params},\39\\{tt})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\7
\&{struct} \\{pattern\_trie} ${}{*}\\{pt}\K\\{init\_pattern\_trie}(\T{256},\39%
\T{128});{}$\7
\&{if} ${}(\\{pt}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\7
\&{struct} \\{pass\_stats} \\{ps};\7
\&{if} ${}(\R\\{read\_patterns}(\\{params},\39\\{pt},\39\\{tt},\39{\AND}%
\\{ps})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\par
\U5.\fi

\N{1}{7}Level range specification.
Besides the command line parameters, the program prompts for the
hyperparameters of the algorithm. The \hyphstartpar{}
and \hyphfinishpar{} specify the range of hyphenation levels covered during the
run. The program can generate patterns up
to level 255.

\Y\B\4\X7:Level range specification\X${}\E{}$\6
\\{printf}(\.{"hyph\_start\ (lowest\ }\)\.{-),\ hyph\_finish\ (hig}\)\.{hest\
hyphenation\ lev}\)\.{el):\ "});\7
\&{size\_t} \\{hyph\_start}${},\39\\{hyph\_finish};{}$\6
\&{int} \\{result};\7
\&{while} (\\{true})\5
${}\{{}$\1\6
${}\\{result}\K\\{scanf}(\.{"\%zu\ \%zu"},\39{\AND}\\{hyph\_start},\39{\AND}%
\\{hyph\_finish});{}$\6
\&{if} ${}(\\{result}\E\T{2}\W\\{hyph\_start}\G\T{1}\W\\{hyph\_start}\Z\T{255}%
\W\\{hyph\_finish}\G\T{1}\W\\{hyph\_finish}\Z\T{254}){}$\5
${}\{{}$\1\6
\&{break};\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\\{printf}(\.{"Error:\ Specify\ 1\ <=}\)\.{\ hyph\_start,\ hyph\_fi}\)\.{nish\
<=\ 255!\ Insert\ }\)\.{again:\ "});\6
\&{while} ${}(\\{getchar}(\,)\I\.{'\\n'}){}$\1\5
;\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{params}\MG\\{hyph\_start}\K(\\{uint8\_t})\\{hyph\_start};{}$\6
\&{if} ${}(\\{hyph\_start}>\\{hyph\_finish}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{hyph\_finish}\K\\{ps}.\\{max\_level};{}$\6
${}\\{printf}(\.{"Warning:\ hyph\_start}\)\.{\ >\ hyph\_finish,\ usin}\)\.{g\
hyph\_finish\ =\ \%u\\n}\)\.{"},\39\\{ps}.\\{max\_level});{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{params}\MG\\{hyph\_finish}\K(\\{uint8\_t})\\{hyph\_finish};{}$\6
\4${}\}{}$\2\par
\U5.\fi

\N{1}{8}Pattern generation loop.
The algorithm runs for each level within the specified range, going from %
\hyphstartpar{} up. The program prompts for
level-specific hyperparameters, generates and prunes the patterns.

\Y\B\4\X8:Pattern generation\X${}\E{}$\6
\&{size\_t} \\{pat\_start}${},\39\\{pat\_finish},\39\\{good\_wt},\39\\{bad%
\_wt},\39\\{thresh};{}$\7
\&{for} (\&{size\_t} \|i${}\K\\{params}\MG\\{hyph\_start};{}$ ${}\|i\Z%
\\{params}\MG\\{hyph\_finish};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{hyph\_level}\K\|i;{}$\6
${}\\{ps}.\\{level\_pattern\_cnt}\K\T{0};{}$\6
\&{if} ${}(\\{params}\MG\\{hyph\_level}>\\{params}\MG\\{hyph\_start}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
\&{if} ${}(\\{params}\MG\\{hyph\_start}\Z\\{ps}.\\{max\_level}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"Warning:\ Largest\ hy}\)\.{phenation\ value\ \%u\ i}\)\.{n\
patterns\ should\ be}\)\.{\ less\ than\ hyph\_star}\)\.{t\\n"},\39\\{ps}.\\{max%
\_level});{}$\6
\4${}\}{}$\2\6
\X9:Level hyperparameters input\X;\6
\X10:Level generation\X;\6
\&{if} ${}(\R\\{delete\_bad\_patterns}(\\{pt})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"total\ of\ \%zu\ patter}\)\.{ns\ at\ hyph\_level\ \%u\\}\)%
\.{n"},\39\\{ps}.\\{level\_pattern\_cnt},\39\\{params}\MG\\{hyph\_level});{}$\6
\4${}\}{}$\2\par
\U5.\fi

\N{1}{9}Level hyperparameters input.
The program again prompts for hyperparameter input. Firstly, it asks for %
\patstartpar{} and \patfinishpar{} that define the
length range of patterns for respective level. The maximum length of a pattern
in \utfpatgen{} is set to 255.
Subsequently, the user is prompted to insert the three weights \goodwtpar{}, %
\badwtpar{} and \threshpar{}. These define the
acceptance criteria for candidate patterns -- in order to accept the pattern
during the ongoing iteration, its number
of {\it good} (supporting) and {\it bad} (contradicting) occurences must make
the following inequality to hold:
$
good * good\_wt - bad * bad\_wt \geq thresh
$
The maximum value for the weights and threshold in \utfpatgen{} is set to 255.

\Y\B\4\X9:Level hyperparameters input\X${}\E{}$\6
\\{printf}(\.{"pat\_start\ (shortest}\)\.{\ -),\ pat\_finish\ (lon}\)\.{gest\
pattern\ explore}\)\.{d):\ "});\6
\&{while} (\\{true})\5
${}\{{}$\1\6
${}\\{result}\K\\{scanf}(\.{"\%zu\ \%zu"},\39{\AND}\\{pat\_start},\39{\AND}%
\\{pat\_finish});{}$\6
\&{if} ${}(\\{result}\E\T{2}\W\\{pat\_start}\G\T{1}\W\\{pat\_finish}\G\T{1}\W%
\\{pat\_start}\Z\\{pat\_finish}\W\\{pat\_start}\Z\T{255}\W\\{pat\_finish}\Z%
\T{255}){}$\5
${}\{{}$\1\6
\&{break};\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\\{printf}(\.{"Error:\ Specify\ 1\ <=}\)\.{\ pat\_start\ <=\ pat\_fi}\)\.{nish\
<=\ 255!\ Insert\ }\)\.{again:\ "});\6
\&{while} ${}(\\{getchar}(\,)\I\.{'\\n'}){}$\1\5
;\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{params}\MG\\{pat\_start}\K(\\{uint8\_t})\\{pat\_start};{}$\6
${}\\{params}\MG\\{pat\_finish}\K(\\{uint8\_t})\\{pat\_finish};{}$\6
\\{printf}(\.{"good\_wt\ (good\ -),\ b}\)\.{ad\_wt\ (bad\ pattern\ w}\)%
\.{eight),\ threshold:\ "});\6
\&{while} (\\{true})\5
${}\{{}$\1\6
${}\\{result}\K\\{scanf}(\.{"\%zu\ \%zu\ \%zu"},\39{\AND}\\{good\_wt},\39{\AND}%
\\{bad\_wt},\39{\AND}\\{thresh});{}$\6
\&{if} ${}(\\{result}\E\T{3}\W\\{good\_wt}\G\T{1}\W\\{bad\_wt}\G\T{1}\W%
\\{thresh}\G\T{1}\W\\{good\_wt}\Z\T{255}\W\\{bad\_wt}\Z\T{255}\W\\{thresh}\Z%
\T{255}){}$\5
${}\{{}$\1\6
\&{break};\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\\{printf}(\.{"Error:\ Specify\ 1\ <=}\)\.{\ good\_wt,\ bad\_wt,\ th}\)%
\.{reshold\ <=\ 255!\ Inse}\)\.{rt\ again:\ "});\6
\&{while} ${}(\\{getchar}(\,)\I\.{'\\n'}){}$\1\5
;\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{params}\MG\\{good\_wt}\K(\\{uint8\_t})\\{good\_wt};{}$\6
${}\\{params}\MG\\{bad\_wt}\K(\\{uint8\_t})\\{bad\_wt};{}$\6
${}\\{params}\MG\\{thresh}\K(\\{uint8\_t})\\{thresh}{}$;\par
\U8.\fi

\N{1}{10}Level generation.
The single pass of \utfpatgen{} at given hyphenation level comprises iterating
through pattern lengths ({\tt pat\_len}
parameter, in ascending order) and dot positions ({\tt pat\_dot}, from the
middle toward the edges) and processing
the dictionary. The algorithm collects supporting and contradicting occurences
and eventually adds new patterns to the
set. It can happen that an iteration is skipped if it is known in advance that
it will not yield any new patterns.

\Y\B\4\X10:Level generation\X${}\E{}$\6
\\{uint8\_t}\\{aux\_dot};\7
\&{bool} \\{more\_this\_level}[\T{256}];\7
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\T{256};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
${}\\{more\_this\_level}[\|i]\K\\{true};{}$\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|j${}\K\\{params}\MG\\{pat\_start};{}$ ${}\|j\Z\\{params}%
\MG\\{pat\_finish};{}$ ${}\|j\PP){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{pat\_len}\K\|j;{}$\6
${}\\{params}\MG\\{pat\_dot}\K\\{params}\MG\\{pat\_len}/\T{2};{}$\6
${}\\{aux\_dot}\K\\{params}\MG\\{pat\_dot}*\T{2};{}$\6
\&{while} ${}(\\{params}\MG\\{pat\_dot}\I\\{params}\MG\\{pat\_len}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{pat\_dot}\K\\{aux\_dot}-\\{params}\MG\\{pat\_dot};{}$\6
${}\\{aux\_dot}\K\\{params}\MG\\{pat\_len}*\T{2}-\\{aux\_dot}-\T{1};{}$\6
\&{if} ${}(\\{more\_this\_level}[\\{params}\MG\\{pat\_dot}]){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{process\_dictionary}(\\{params},\39\\{tt},\39\\{pt},\39{\AND}%
\\{ps})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\6
${}\\{more\_this\_level}[\\{params}\MG\\{pat\_dot}]\K\\{ps}.\\{more\_to%
\_come};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{255};{}$ ${}\|i>\T{0};{}$ ${}\|i\MM){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{more\_this\_level}[\|i-\T{1}]){}$\5
${}\{{}$\1\6
${}\\{more\_this\_level}[\|i]\K\\{false};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U8.\fi

\N{1}{11}Final pass.
If the user wishes, the dictionary is traversed one last time and hyphenated
according to found patterns. The output is
stored in file 'pattmp.X'.

\Y\B\4\X11:Final pass\X${}\E{}$\6
\&{if} ${}(\R\\{output\_patterns}(\\{pt},\39\\{tt},\39\\{params}\MG\\{output%
\_file})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\7
\&{char} \|c;\6
\&{bool} \\{output}${}\K\\{false};{}$\7
\\{printf}(\.{"hyphenate\ word\ list}\)\.{?\ (y/n):\ "});\6
\&{if} ${}(\\{scanf}(\.{"\ \%c"},\39{\AND}\|c)<\T{1}){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\6
\&{if} ${}(\|c\E\.{'y'}\V\|c\E\.{'Y'}){}$\5
${}\{{}$\1\6
${}\\{output}\K\\{true};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{hyphenate\_dictionary}(\\{params},\39\\{tt},\39\\{pt},\39%
\\{output},\39{\AND}\\{ps})){}$\5
${}\{{}$\1\6
\\{destroy\_params}(\\{params});\6
\\{destroy\_tr\_table}(\\{tt});\6
\\{destroy\_pattern\_trie}(\\{pt});\6
\&{return} \.{EXIT\_FAILURE};\6
\4${}\}{}$\2\par
\U5.\fi

\N{0}{12}Implementation details.
In this section we focus on the building blocks, each subsection devoted to one
particular aspect of the algorithm.

\fi

\N{1}{13}IO procedures.
Methods in this subsection stand on the interface between \utfpatgen{} and the
user, together with the {\tt main}
method. {\tt parse\_inputs} attempts to open the 4 files provided as inputs
into streams and save them for later
use. {\tt read\_line} is a utility to simplify reading from these streams that
stores the information read into
provided buffer. The {\tt print\_help} and {\tt print\_version} methods print
out the desired information if the
user does not wish to proceed to pattern generation.

\fi

\M{14}parse\_input.
Attempts to open the 4 files provided as inputs into streams and save them for
later use. Returns a boolean indicating
whether all files have been opened successfully.

\Y\B\1\1\&{bool} \\{parse\_input}(\&{char} ${}{*}\\{argv}[\,],\39{}$\&{int} %
\\{argc}${},\39{}$\&{struct} \\{params} ${}{*}\\{params})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{argc}\I\T{5}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"utfpatgen\ need\ exac}\)\.{tly\ 4\ arguments.%
\\nTr}\)\.{y\ `utfpatgen\ --help`}\)\.{\ for\ more\ informatio}\)\.{n.%
\\n"});{}$\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{dictionary\_file}\K\\{fopen}(\\{argv}[\T{1}],\39\.{"rb"});{}$%
\7
\&{if} ${}(\\{dictionary\_file}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Could\ not\ open\ dict}\)\.{ionary\ file\ '%
\%s'.\\n"},\39\\{argv}[\T{1}]);{}$\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{pattern\_file}\K\\{fopen}(\\{argv}[\T{2}],\39\.{"rb"});{}$\7
\&{if} ${}(\\{pattern\_file}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Could\ not\ open\ patt}\)\.{ern\ file\ '\%s'.%
\\n"},\39\\{argv}[\T{2}]);{}$\6
\\{fclose}(\\{dictionary\_file});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{output\_file}\K\\{fopen}(\\{argv}[\T{3}],\39\.{"wb"});{}$\7
\&{if} ${}(\\{output\_file}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Could\ not\ open\ outp}\)\.{ut\ file\ '\%s'.%
\\n"},\39\\{argv}[\T{3}]);{}$\6
\\{fclose}(\\{dictionary\_file});\6
\\{fclose}(\\{pattern\_file});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{translate\_file}\K\\{fopen}(\\{argv}[\T{4}],\39\.{"rb"});{}$\7
\&{if} ${}(\\{translate\_file}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Could\ not\ open\ tran}\)\.{slate\ file\ '%
\%s'.\\n"},\39\\{argv}[\T{4}]);{}$\6
\\{fclose}(\\{dictionary\_file});\6
\\{fclose}(\\{pattern\_file});\6
\\{fclose}(\\{output\_file});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{params}\MG\\{dictionary\_file}\K\\{dictionary\_file};{}$\6
${}\\{params}\MG\\{pattern\_file}\K\\{pattern\_file};{}$\6
${}\\{params}\MG\\{output\_file}\K\\{output\_file};{}$\6
${}\\{params}\MG\\{translate\_file}\K\\{translate\_file};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{15}read\_line.
Reads a line from given stream into a string buffer. If end of file is reached,
it sets {\tt eof} flag of the buffer.
Returns a boolean indicating whether the line has been read successfully.

\Y\B\1\1\&{bool} \\{read\_line}(\&{FILE} ${}{*}\\{stream},\39{}$\&{struct} %
\\{string\_buffer} ${}{*}\\{buf})\2\2{}$\6
${}\{{}$\1\6
\\{reset\_buffer}(\\{buf});\7
\&{char} \|c;\7
\&{while} ${}((\|c\K\\{fgetc}(\\{stream}))\I\.{EOF}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{buf}\MG\\{size}\G\\{buf}\MG\\{capacity}-\T{1}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{resize\_buffer}(\\{buf},\39\T{2}*\\{buf}\MG\\{capacity})\E%
\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\|c\E\.{'\\n'}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{buf}\MG\\{size}>\T{0})\W(\\{buf}\MG\\{data}[\\{buf}\MG\\{size}-%
\T{1}]\E\.{'\\r'})){}$\5
${}\{{}$\SHC{ Windows /r/n end of line }\1\6
${}\\{buf}\MG\\{size}\MRL{-{\K}}\T{1};{}$\6
\4${}\}{}$\2\6
\&{break};\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{data}[\\{buf}\MG\\{size}]\K\|c;{}$\6
${}\\{buf}\MG\\{size}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{data}[\\{buf}\MG\\{size}]\K\.{'\\0'};{}$\6
\&{if} ${}(\|c\E\.{EOF}){}$\5
${}\{{}$\1\6
${}\\{buf}\MG\\{eof}\K\\{true};{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{16}print\_help.
Prints out a help message to the standard output.

\Y\B\1\1\&{void} \\{print\_help}(\,)\2\2\6
${}\{{}$\1\6
\\{printf}(\.{"Usage:\ utfpatgen\ [O}\)\.{PTION]...\ DICTIONARY}\)\.{\ PATTERNS%
\ OUTPUT\ TRA}\)\.{NSLATE\\n"});\6
\\{printf}(\.{"\\tGenerate\ the\ OUTP}\)\.{UT\ hyphenation\ file\ }\)\.{for\
use\ with\ TeX\\n"});\6
\\{printf}(\.{"\\tfrom\ the\ DICTIONA}\)\.{RY,\ PATTERNS,\ and\ TR}\)\.{ANSLATE%
\ files.\\n"});\6
\\{printf}(\.{"\\n--help\ \ \ \ \ \ \ \ pri}\)\.{nt\ this\ help\ and\ exi}\)%
\.{t\\n"});\6
\\{printf}(\.{"--version\ \ \ \ \ outpu}\)\.{t\ version\ informatio}\)\.{n\ and%
\ exit\\n"});\6
\4${}\}{}$\2\par
\fi

\M{17}print\_version.
Prints out the version number of \utfpatgen{} to the standard output.

\Y\B\1\1\&{void} \\{print\_version}(\,)\2\2\6
${}\{{}$\1\6
${}\\{printf}(\.{"This\ is\ utfpatgen\ v}\)\.{ersion\ \%s\\n"},\39\.{UTFPATGEN%
\_VERSION});{}$\6
\4${}\}{}$\2\par
\fi

\N{1}{18}Translate file processing.
Once the translate file has been successfully opened and its stream pointer
stored, it is read line by line into
translate table. The purpose of the translate file is to list all the
characters that occur in the dictionary, together
with their uppercase variants. Later during the pattern generation, all of
these variants are treated as though they
were the same character. Besides, the first line of the file may redefine the
language-specific parameters
\lefthyphenminpar, \righthyphenminpar, and the symbols used for marking hyphens
in the dictionary. The format of the
translate file follows the same pattern as required by \patgen:

\item{$\bullet$} first line (optional): 'LLRR BMG', where 'LL' is the value of %
\lefthyphenminpar, 'RR' the value of
\righthyphenminpar, 'B' the symbol for {\tt BAD\_HYF}, 'M' the symbol for {\tt
MISS\_HYF}, and 'G' the
symbol for {\tt GOOD\_HYF}. Should any of these parameters stay blank, the
default is used:
\lefthyphenminpar$=2$, \righthyphenminpar$=3$, {\tt BAD\_HYF} '.', {\tt MISS%
\_HYF} '-',
{\tt GOOD\_HYF} '*'.
\item{$\bullet$} consequent lines: '$\_X\_Y_1\_...Y_n\_\_$', where '$X$' is a
lowercase letter, '$Y_k$' an arbitrary
(even 0) number of upper-case variants of '$X$', and '$\_$' a delimiter,
usually space.

Note that this format allows two-digit values of \lefthyphenminpar{} and %
\righthyphenminpar{} at most and one-byte
characters for hyphen symbols. Practically, this is not a problem.

The translate file must be provided as an input to \utfpatgen, but may be left
empty. In that case, the default values
for parameters ASCII character mapping are used.

\fi

\M{19}read\_translate.
Reads the contents of the translate file and parses its lines into a translate
table. It includes parsing the optional
header line and storing the parameters obtained. If the file is empty, the
method fetches a default ASCII mapping.
Returns a boolean indicating whether the file has been read successfully.

\Y\B\1\1\&{bool} \\{read\_translate}(\&{struct} \\{params} ${}{*}\\{params},%
\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt})\2\2{}$\6
${}\{{}$\1\6
${}\\{rewind}(\\{params}\MG\\{translate\_file});{}$\7
\&{struct} \\{string\_buffer} ${}{*}\\{buf}\K\\{init\_buffer}(\T{64});{}$\7
\&{if} ${}(\\{buf}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{trie} ${}{*}\\{helper\_trie}\K\\{init\_trie}(\T{256});{}$\7
\&{if} ${}(\\{helper\_trie}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{read\_line}(\\{params}\MG\\{translate\_file},\39\\{buf})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{buf}\MG\\{eof}){}$\5
${}\{{}$\1\6
\&{bool} \\{default\_mapping}${}\K\\{default\_ascii\_mapping}(\\{tt},\39%
\\{helper\_trie});{}$\7
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{default\_mapping};\6
\4${}\}{}$\2\7
\&{size\_t} \\{out\_index}${},\39\\{letter\_index};{}$\7
${}\\{letter\_index}\K\\{tt}\MG\\{letter\_count}+\T{1};{}$\6
\&{if} ${}(\\{letter\_index}\G\\{tt}\MG\\{letter\_capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\\{tt}\MG\\{letter\_capacity}*\T{2};{}$\6
\&{size\_t} ${}{*}\\{new\_array}\K\\{realloc}(\\{tt}\MG\\{index\_to\_alphabet},%
\39\\{new\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_array}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}\K\\{new\_array};{}$\6
${}\\{tt}\MG\\{letter\_capacity}\K\\{new\_capacity};{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]\K\\{tt}\MG\\{alphabet}%
\MG\\{size};{}$\6
\&{if} ${}(\R\\{insert\_pattern}(\\{tt}\MG\\{mapping},\39{}$(\&{const} %
\&{char}[\,])\,${}\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\},\39{\AND}\\{out\_index},%
\39\\{helper\_trie})\V\R\\{append\_string}(\\{tt}\MG\\{alphabet},\39{}$(%
\&{const} \&{char}[\,])\,${}\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\MG\\{aux}[\\{out\_index}]\K\\{letter\_index};{}$\6
${}\\{tt}\MG\\{letter\_count}\PP;{}$\7
\&{bool} \\{first\_line}${}\K\\{true};{}$\7
\&{while} ${}(\R\\{buf}\MG\\{eof}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{first\_line}\W\\{parse\_header}(\\{buf},\39\\{params})){}$\5
${}\{{}$\SHC{ header parsed successfully }\6
\,${}\}{}$\6
\&{else} \&{if} ${}(\R\\{parse\_letters}(\\{buf},\39\\{tt},\39\\{helper%
\_trie})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{first\_line}\K\\{false};{}$\6
\\{reset\_buffer}(\\{buf});\6
\&{if} ${}(\R\\{read\_line}(\\{params}\MG\\{translate\_file},\39\\{buf})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
${}\\{printf}(\.{"left\_hyphen\_min\ =\ \%}\)\.{u,\ right\_hyphen\_min\ }\)\.{=%
\ \%u,\ \%zu\ letters\\n"},\39\\{params}\MG\\{left\_hyphen\_min},\39\\{params}%
\MG\\{right\_hyphen\_min},\39\\{tt}\MG\\{letter\_count});{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{20}is\_integer.
Returns a boolean indicating whether the given character is an ASCII numeric
literal.

\Y\B\1\1\&{bool} \\{is\_integer}(\&{char} \|c)\2\2\6
${}\{{}$\1\6
\&{return} ${}(\|c\G\.{'0'}\W\|c\Z\.{'9'});{}$\6
\4${}\}{}$\2\par
\fi

\M{21}is\_space.
Returns a boolean indicating whether the given character is an ASCII whitespace
(0x20).

\Y\B\1\1\&{bool} \\{is\_space}(\&{char} \|c)\2\2\6
${}\{{}$\1\6
\&{return} ${}(\|c\E\.{'\ '});{}$\6
\4${}\}{}$\2\par
\fi

\M{22}parse\_two\_digit.
Attempts to parse 2-character sequence from a buffer into 2-digit number.
Returns a boolean indicating whether the
sequence is representing a number and parsing was finished successfully.

\Y\B\1\1\&{bool} \\{parse\_two\_digit}(\&{struct} \\{string\_buffer} ${}{*}%
\\{buf},\39{}$\&{size\_t} \\{pos}${},\39\\{int8\_t}*\\{out})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{pos}+\T{1}\G\\{buf}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{char} \\{c1}${}\K\\{buf}\MG\\{data}[\\{pos}];{}$\6
\&{char} \\{c2}${}\K\\{buf}\MG\\{data}[\\{pos}+\T{1}];{}$\7
\&{if} ${}(\\{is\_space}(\\{c1})\W\\{is\_space}(\\{c2})){}$\5
${}\{{}$\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\6
\&{if} (\\{is\_space}(\\{c1}))\5
${}\{{}$\1\6
${}\\{c1}\K\.{'0'};{}$\6
\4${}\}{}$\2\6
\&{if} (\\{is\_space}(\\{c2}))\5
${}\{{}$\1\6
${}\\{c2}\K\.{'0'};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{is\_integer}(\\{c1})\V\R\\{is\_integer}(\\{c2})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}{*}\\{out}\K(\\{c1}-\.{'0'})*\T{10}+(\\{c2}-\.{'0'});{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{23}parse\_header.
Attempts to parse the text stored in a buffer as though it was the header of a
translate file. If successful, the
values of parameters are stored in {\tt params}. Return value indicates whether
the parsing succeeded.

\Y\B\1\1\&{bool} \\{parse\_header}(\&{struct} \\{string\_buffer} ${}{*}\\{buf},%
\39{}$\&{struct} \\{params} ${}{*}\\{params})\2\2{}$\6
${}\{{}$\1\6
${}\\{int8\_t}\\{val}\K{-}\T{1};{}$\6
\&{if} ${}(\R\\{parse\_two\_digit}(\\{buf},\39\T{0},\39{\AND}\\{val})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{val}\I{-}\T{1}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{left\_hyphen\_min}\K\\{val};{}$\6
\4${}\}{}$\2\6
${}\\{val}\K{-}\T{1};{}$\6
\&{if} ${}(\R\\{parse\_two\_digit}(\\{buf},\39\T{2},\39{\AND}\\{val})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{val}\I{-}\T{1}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{right\_hyphen\_min}\K\\{val};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{buf}\MG\\{size}\G\T{6}\W\R\\{is\_space}(\\{buf}\MG\\{data}[%
\T{5}])){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{bad\_hyphen}\K\\{buf}\MG\\{data}[\T{5}];{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{buf}\MG\\{size}\G\T{7}\W\R\\{is\_space}(\\{buf}\MG\\{data}[%
\T{6}])){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{missed\_hyphen}\K\\{buf}\MG\\{data}[\T{6}];{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{buf}\MG\\{size}\G\T{8}\W\R\\{is\_space}(\\{buf}\MG\\{data}[%
\T{7}])){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{good\_hyphen}\K\\{buf}\MG\\{data}[\T{7}];{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{24}parse\_letters.
Attempts to parse the text stored in a buffer as though it was a non-header
line of a translate file. If successful,
the respective letter and its uppercase variants are stored in the translation
table. If any of the letters was already
in the table, the method fails. Return values indicates the success of parsing.

\Y\B\1\1\&{bool} \\{parse\_letters}(\&{struct} \\{string\_buffer} ${}{*}%
\\{buf},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} %
\\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{buf}\MG\\{size}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Empty\ line\ in\ trans}\)\.{late\ file%
\\n"});{}$\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{char} \\{separator}${}\K\\{buf}\MG\\{data}[\T{0}];{}$\7
\&{if} ${}(\\{buf}\MG\\{size}>\T{1}\W\\{buf}\MG\\{data}[\T{1}]\E%
\\{separator}){}$\5
${}\{{}$\SHC{ a comment }\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\&{size\_t} \\{letter\_index}${}\K\\{tt}\MG\\{letter\_count}+\T{1};{}$\6
\&{size\_t} \\{out\_index};\6
\&{struct} \\{string\_buffer} ${}{*}\\{letter}\K\\{init\_buffer}(\T{4});{}$\7
\&{if} ${}(\\{letter}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{bool} \\{lower}${}\K\\{true};{}$\7
\&{if} ${}(\R\\{append\_char}(\\{buf},\39\\{separator})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{1};{}$ ${}\|i<\\{buf}\MG\\{size};{}$ ${}\|i%
\PP){}$\5
${}\{{}$\1\6
\&{char} \|c${}\K\\{buf}\MG\\{data}[\|i];{}$\7
\&{if} ${}(\|c\E\\{separator}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{letter}\MG\\{size}\E\T{0}){}$\5
${}\{{}$\1\6
\&{break};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\.{'\\0'})\V\R\\{insert\_pattern}(%
\\{tt}\MG\\{mapping},\39\\{letter}\MG\\{data},\39{\AND}\\{out\_index},\39%
\\{helper\_trie})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\MG\\{aux}[\\{out\_index}]\K\\{letter\_index};{}$\6
\&{if} (\\{lower})\5
${}\{{}$\1\6
\&{if} ${}(\\{letter\_index}\G\\{tt}\MG\\{letter\_capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\\{tt}\MG\\{letter\_capacity}*\T{2};{}$\6
\&{size\_t} ${}{*}\\{new\_array}\K\\{realloc}(\\{tt}\MG\\{index\_to\_alphabet},%
\39\\{new\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_array}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}\K\\{new\_array};{}$\6
${}\\{tt}\MG\\{letter\_capacity}\K\\{new\_capacity};{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]\K\\{tt}\MG\\{alphabet}%
\MG\\{size};{}$\6
\&{if} ${}(\R\\{append\_string}(\\{tt}\MG\\{alphabet},\39\\{letter}\MG%
\\{data})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{letter\_count}\PP;{}$\6
${}\\{lower}\K\\{false};{}$\6
\4${}\}{}$\2\6
\\{reset\_buffer}(\\{letter});\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\|c)){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{25}default\_ascii\_mapping.
Fetches default ASCII character mapping into the translate table. Return value
indicate the success of fetching.

\Y\B\1\1\&{bool} \\{default\_ascii\_mapping}(\&{struct} \\{translate\_table}
${}{*}\\{tt},\39{}$\&{struct} \\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{out\_index};\6
\&{size\_t} \\{letter\_index};\6
\&{char} \\{upper};\7
\&{for} (\&{char} \|c${}\K\.{'a'};{}$ ${}\|c\Z\.{'z'};{}$ ${}\|c\PP){}$\5
${}\{{}$\1\6
${}\\{letter\_index}\K\\{tt}\MG\\{letter\_count}+\T{1};{}$\6
${}\\{upper}\K\|c-(\.{'a'}-\.{'A'});{}$\6
\&{if} ${}(\\{letter\_index}\G\\{tt}\MG\\{letter\_capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\\{tt}\MG\\{letter\_capacity}*\T{2};{}$\6
\&{size\_t} ${}{*}\\{new\_array}\K\\{realloc}(\\{tt}\MG\\{index\_to\_alphabet},%
\39\\{new\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_array}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}\K\\{new\_array};{}$\6
${}\\{tt}\MG\\{letter\_capacity}\K\\{new\_capacity};{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]\K\\{tt}\MG\\{alphabet}%
\MG\\{size};{}$\6
\&{if} ${}(\R\\{insert\_pattern}(\\{tt}\MG\\{mapping},\39{}$(\&{const} %
\&{char}[\,])\,${}\{\|c,\39\.{'\\0'}\},\39{\AND}\\{out\_index},\39\\{helper%
\_trie})\V\R\\{append\_string}(\\{tt}\MG\\{alphabet},\39{}$(\&{const} \&{char}[%
\,])\,${}\{\|c,\39\.{'\\0'}\})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\MG\\{aux}[\\{out\_index}]\K\\{letter\_index};{}$\6
\&{if} ${}(\R\\{insert\_pattern}(\\{tt}\MG\\{mapping},\39{}$(\&{const} %
\&{char}[\,])\,${}\{\\{upper},\39\.{'\\0'}\},\39{\AND}\\{out\_index},\39%
\\{helper\_trie})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\MG\\{aux}[\\{out\_index}]\K\\{letter\_index};{}$\6
${}\\{tt}\MG\\{letter\_count}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{letter\_index}\K\\{tt}\MG\\{letter\_count}+\T{1};{}$\6
\&{if} ${}(\\{letter\_index}\G\\{tt}\MG\\{letter\_capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\\{tt}\MG\\{letter\_capacity}*\T{2};{}$\6
\&{size\_t} ${}{*}\\{new\_array}\K\\{realloc}(\\{tt}\MG\\{index\_to\_alphabet},%
\39\\{new\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_array}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}\K\\{new\_array};{}$\6
${}\\{tt}\MG\\{letter\_capacity}\K\\{new\_capacity};{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]\K\\{tt}\MG\\{alphabet}%
\MG\\{size};{}$\6
\&{if} ${}(\R\\{insert\_pattern}(\\{tt}\MG\\{mapping},\39{}$(\&{const} %
\&{char}[\,])\,${}\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\},\39{\AND}\\{out\_index},%
\39\\{helper\_trie})\V\R\\{append\_string}(\\{tt}\MG\\{alphabet},\39{}$(%
\&{const} \&{char}[\,])\,${}\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\MG\\{aux}[\\{out\_index}]\K\\{letter\_index};{}$\6
${}\\{tt}\MG\\{letter\_count}\PP;{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{26}Pattern file processing.
The user can provide \utfpatgen{} with initial set of patterns to work with.
Similarly to the translate file, the pattern
file must be given as input, but may be left empty. Each line of the file
represents one pattern, with following format
required:
{\narrower
$<hyph. level><character><hyph. level>\dots<hyph. level>$
\par}
Zero levels are omitted implicitly. A special character ('.' in \patgen, \tt
EDGE\_OF\_WORD '0xff' in \utfpatgen)
denotes edges of the word (it can be only used as the first or last character
of a pattern). The original \patgen
program that supports only levels up to 9 represents them simply as ASCII
numeric literals '0' to '9'. On the other
hand, \utfpatgen{} expects each byte representing a level preceded with a
special {\tt HYPHEN\_FLAG} byte of hexadecimal
value '0xfe', so the range is extended up to 253. The '0xfe' and '0xff' bytes
are not used by any UTF-8 character by
design.

\fi

\M{27}read\_patterns.
Iterates over the pattern file and reads its entries into a pattern trie.
Return value indicates whether the whole file
has been read and parsed successfully.

\Y\B\1\1\&{bool} \\{read\_patterns}(\&{struct} \\{params} ${}{*}\\{params},%
\39{}$\&{struct} \\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} \\{translate%
\_table} ${}{*}\\{tt},\39{}$\&{struct} \\{pass\_stats} ${}{*}\\{ps})\2\2{}$\6
${}\{{}$\1\6
${}\\{ps}\MG\\{level\_pattern\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{max\_level}\K\T{0};{}$\7
\&{struct} \\{string\_buffer} ${}{*}\\{buf}\K\\{init\_buffer}(\T{16});{}$\7
\&{if} ${}(\\{buf}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{eof}\K\\{false};{}$\7
\&{struct} \\{pattern} ${}{*}\\{pat}\K\\{init\_pattern}(\T{16});{}$\7
\&{if} ${}(\\{pat}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{trie} ${}{*}\\{helper\_trie}\K\\{init\_trie}(\T{256});{}$\7
\&{if} ${}(\\{helper\_trie}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_pattern}(\\{pat});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{while} ${}(\R\\{buf}\MG\\{eof}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{read\_line}(\\{params}\MG\\{pattern\_file},\39\\{buf})\V\R%
\\{parse\_pattern}(\\{buf},\39\\{pat},\39\\{tt})\V\R\\{insert\_new\_pattern}(%
\\{pat},\39\\{pt},\39\\{ps},\39\\{helper\_trie})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_pattern}(\\{pat});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\%zu\ patterns\ read\ i}\)\.{n\\n"},\39\\{ps}\MG\\{level%
\_pattern\_cnt});{}$\6
${}\\{printf}(\.{"pattern\ trie\ has\ \%z}\)\.{u\ nodes,\ trie\_max\ =\ }\)\.{%
\%zu,\ \%zu\ outputs\\n"},\39\\{pt}\MG\|t\MG\\{occupied},\39\\{pt}\MG\|t\MG%
\\{node\_max},\39\\{pt}\MG\\{ops}\MG\\{count});{}$\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_pattern}(\\{pat});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{28}parse\_pattern.
Processes the pattern text from buffer, translates it to lowercase and inserts
it into a pattern structure. Return
value indicates the success of translation and parsing.

\Y\B\1\1\&{bool} \\{parse\_pattern}(\&{struct} \\{string\_buffer} ${}{*}%
\\{buf},\39{}$\&{struct} \\{pattern} ${}{*}\\{out\_pattern},\39{}$\&{struct} %
\\{translate\_table} ${}{*}\\{tt})\2\2{}$\6
${}\{{}$\1\6
\\{reset\_pattern}(\\{out\_pattern});\7
\&{char} \|c;\6
\&{bool} \\{next\_hyphen}${}\K\\{false};{}$\6
\&{struct} \\{string\_buffer} ${}{*}\\{letter}\K\\{init\_buffer}(\T{4});{}$\7
\&{if} ${}(\\{letter}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{buf}\MG\\{size};{}$ ${}\|i%
\PP){}$\5
${}\{{}$\1\6
${}\|c\K\\{buf}\MG\\{data}[\|i];{}$\6
\&{if} ${}(\|c\E\.{EDGE\_OF\_WORD}\W\|i\I\T{0}\W\|i\I\\{buf}\MG\\{size}-%
\T{1}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Edge\ of\ word\ found\ }\)\.{inside\ a\
pattern.\\n"});{}$\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} (\\{next\_hyphen})\5
${}\{{}$\1\6
\&{if} ${}(\R\\{set\_hyphen}(\\{out\_pattern},\39\\{out\_pattern}\MG\\{size},%
\39(\\{uint8\_t})\|c)){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{next\_hyphen}\K\\{false};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{is\_utf\_start\_byte}(\|c)\W\\{letter}\MG\\{size}>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\.{'\\0'})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{letter\_index}${}\K\\{get\_letter\_index}(\\{tt},\39\\{letter}%
\MG\\{data});{}$\7
\&{if} ${}(\\{letter\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Unknown\ letter\ \%s\ f}\)\.{ound\ in\ a\
pattern.\\n}\)\.{"},\39\\{letter}\MG\\{data});{}$\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{convert\_index\_to\_pattern}(\\{letter\_index},\39\\{out%
\_pattern})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\\{reset\_buffer}(\\{letter});\6
\4${}\}{}$\2\6
\&{if} ${}(\|c\E\.{HYPHEN\_FLAG}){}$\5
${}\{{}$\1\6
${}\\{next\_hyphen}\K\\{true};{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\R\\{append\_char}(\\{letter},\39\|c)){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} (\\{next\_hyphen})\5
${}\{{}$\1\6
\&{if} ${}(\R\\{set\_hyphen}(\\{out\_pattern},\39\\{out\_pattern}\MG\\{size},%
\39(\\{uint8\_t})\|c)){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{letter}\MG\\{size}>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\.{'\\0'})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{letter\_index}${}\K\\{get\_letter\_index}(\\{tt},\39\\{letter}%
\MG\\{data});{}$\7
\&{if} ${}(\\{letter\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Unknown\ letter\ \%s\ f}\)\.{ound\ in\ a\
pattern.\\n}\)\.{"},\39\\{letter}\MG\\{data});{}$\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{convert\_index\_to\_pattern}(\\{letter\_index},\39\\{out%
\_pattern})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{29}insert\_new\_pattern.
Inserts the pattern into pattern trie and collects statistics along the way.
Return value indicates the success of
insertion.

\Y\B\1\1\&{bool} \\{insert\_new\_pattern}(\&{struct} \\{pattern} ${}{*}\\{pat},%
\39{}$\&{struct} \\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} \\{pass%
\_stats} ${}{*}\\{ps},\39{}$\&{struct} \\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{hyphenation\_value}${},\39\\{node};{}$\6
\&{size\_t} \\{current\_len}${}\K\T{0};{}$\7
\&{if} ${}(\R\\{insert\_pattern}(\\{pt}\MG\|t,\39\\{pat}\MG\\{text},\39{\AND}%
\\{node},\39\\{helper\_trie})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{pat}\MG\\{size};{}$ ${}\|i%
\PP){}$\5
${}\{{}$\1\6
${}\\{hyphenation\_value}\K\\{get\_hyphen}(\\{pat},\39\|i);{}$\6
\&{if} ${}(\\{hyphenation\_value}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{ps}\MG\\{level\_pattern\_cnt}\PP;{}$\6
\&{if} ${}(\R\\{set\_output}(\\{pt},\39\\{node},\39\\{hyphenation\_value},\39%
\\{current\_len})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{hyphenation\_value}>\\{ps}\MG\\{max\_level}){}$\5
${}\{{}$\1\6
${}\\{ps}\MG\\{max\_level}\K\\{hyphenation\_value};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}((\\{uint8\_t})\\{pat}\MG\\{text}[\|i]\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_len}\PP;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{hyphenation\_value}\K\\{get\_hyphen}(\\{pat},\39\\{pat}\MG\\{size});{}$\6
\&{if} ${}(\\{hyphenation\_value}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{ps}\MG\\{level\_pattern\_cnt}\PP;{}$\6
\&{if} ${}(\R\\{set\_output}(\\{pt},\39\\{node},\39\\{hyphenation\_value},\39%
\\{current\_len})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{30}Dictionary file processing.
The dictionary is the main source of data for pattern generation. The file
indeed has to be provided as input, and
though no error is raised when it is empty, such case does not make much sense.
Each line in the file represents single
hyphenated word, with hyphens marked using {\tt GOOD\_HYF}, {\tt MISS\_HYF},
and {\tt BAD\_HYF}. Furthermore,
both the whole word and separate hyphens can be weighted by preceding the with
{\tt HYPHEN\_FLAG} and a value.
Similarly to the pattern file, the possible range of weights is increased to
253. If the hyphen weight is omitted, word
weight is used. If the word weight is omitted, \utfpatgen{} uses the default
value 1. The lines of the dictionary file
thus look like this:

{\narrower
$<word weigth><character><hyphen weight><hyphen><character>\dots<character>$
\par}

Since some patterns may be tied to the edges of the word, special byte symbol {%
\tt EDGE\_OF\_WORD} was introduced
that marks the edges. Hexadecimal value of the symbol is '0xff' that is not
used by the UTF-8 encoding.

\fi

\M{31}process\_dictionary.
Reads and parses the dictionary, and generates new patterns afterwards.
Statistics of the pass are printed out at the
end of the method. Return value indicates the success of reading, parsing, and
pattern generation.

\Y\B\1\1\&{bool} \\{process\_dictionary}(\&{struct} \\{params} ${}{*}%
\\{params},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} %
\\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} \\{pass\_stats} ${}{*}\\{ps})\2%
\2{}$\6
${}\{{}$\1\6
${}\\{ps}\MG\\{good\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{bad\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{miss\_cnt}\K\T{0};{}$\6
${}\\{params}\MG\\{word\_weight}\K\T{1};{}$\6
\&{if} ${}(\\{params}\MG\\{hyph\_level}\MOD\T{2}\E\T{1}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{good\_dot}\K\.{MISS\_HYF};{}$\6
${}\\{params}\MG\\{bad\_dot}\K\.{NO\_HYF};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{params}\MG\\{good\_dot}\K\.{BAD\_HYF};{}$\6
${}\\{params}\MG\\{bad\_dot}\K\.{GOOD\_HYF};{}$\6
\4${}\}{}$\2\7
\&{struct} \\{count\_trie} ${}{*}\\{ct}\K\\{init\_count\_trie}(\T{256},\39%
\T{256});{}$\7
\&{if} ${}(\\{ct}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"processing\ dictiona}\)\.{ry\ with\ pat\_len\ =\ \%u}\)\.{,\
pat\_dot\ =\ \%u\\n"},\39\\{params}\MG\\{pat\_len},\39\\{params}\MG\\{pat%
\_dot});{}$\6
\&{if} ${}(\R\\{process\_all\_words}(\\{params},\39\\{tt},\39\\{pt},\39\\{ps},%
\39\\{ct})){}$\5
${}\{{}$\1\6
\\{destroy\_count\_trie}(\\{ct});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\\n\%zu\ good,\ \%zu\ bad}\)\.{,\ \%zu\ missed\\n"},\39\\{ps}%
\MG\\{good\_cnt},\39\\{ps}\MG\\{bad\_cnt},\39\\{ps}\MG\\{miss\_cnt});{}$\6
\&{if} ${}(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG\\{miss\_cnt}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%.2f\ \%\%,\ \%.2f\ \%\%,\ \%}\)\.{.2f\ \%\%\\n"},\39(%
\T{100}*(\&{float})\,\\{ps}\MG\\{good\_cnt}/(\&{float})(\\{ps}\MG\\{good\_cnt}+%
\\{ps}\MG\\{miss\_cnt})),\39(\T{100}*(\&{float})\,\\{ps}\MG\\{bad\_cnt}/(%
\&{float})(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG\\{miss\_cnt})),\39(\T{100}*(%
\&{float})\,\\{ps}\MG\\{miss\_cnt}/(\&{float})(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG%
\\{miss\_cnt})));{}$\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\%zu\ patterns,\ \%zu\ n}\)\.{odes\ in\ count\ trie,\ }\)%
\.{triec\_max\ =\ \%zu\\n"},\39\\{ct}\MG\|t\MG\\{pattern\_count},\39\\{ct}\MG%
\|t\MG\\{occupied},\39\\{ct}\MG\|t\MG\\{node\_max});{}$\6
\&{if} ${}(\R\\{collect\_count\_trie}(\\{ct},\39\\{pt},\39\\{params},\39%
\\{ps})){}$\5
${}\{{}$\1\6
\\{destroy\_count\_trie}(\\{ct});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\\{destroy\_count\_trie}(\\{ct});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{32}process\_all\_words.
Iterates over the lines of the dictionary files and parses them to the count
trie. Return value indicates whether the
whole file has been processed successfully.

\Y\B\1\1\&{bool} \\{process\_all\_words}(\&{struct} \\{params} ${}{*}%
\\{params},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} %
\\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} \\{pass\_stats} ${}{*}\\{ps},%
\39{}$\&{struct} \\{count\_trie} ${}{*}\\{ct})\2\2{}$\6
${}\{{}$\1\6
${}\\{rewind}(\\{params}\MG\\{dictionary\_file});{}$\6
${}\\{uint8\_t}\\{dot\_min}\K\\{params}\MG\\{pat\_dot};{}$\6
${}\\{uint8\_t}\\{dot\_max}\K\\{params}\MG\\{pat\_len}-\\{params}\MG\\{pat%
\_dot};{}$\6
\&{if} ${}(\\{dot\_min}<\\{params}\MG\\{left\_hyphen\_min}+\T{1}){}$\5
${}\{{}$\1\6
${}\\{dot\_min}\K\\{params}\MG\\{left\_hyphen\_min}+\T{1};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{dot\_max}<\\{params}\MG\\{right\_hyphen\_min}+\T{1}){}$\5
${}\{{}$\1\6
${}\\{dot\_max}\K\\{params}\MG\\{right\_hyphen\_min}+\T{1};{}$\6
\4${}\}{}$\2\7
\&{size\_t} \\{dot\_len}${}\K\\{dot\_min}+\\{dot\_max};{}$\6
\&{struct} \\{string\_buffer} ${}{*}\\{buf}\K\\{init\_buffer}(\T{64});{}$\7
\&{if} ${}(\\{buf}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{eof}\K\\{false};{}$\7
\&{struct} \\{word} ${}{*}\\{word}\K\\{init\_word}(\T{16});{}$\7
\&{if} ${}(\\{word}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{trie} ${}{*}\\{helper\_trie}\K\\{init\_trie}(\T{256});{}$\7
\&{if} ${}(\\{helper\_trie}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_word}(\\{word});\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{while} ${}(\R\\{buf}\MG\\{eof}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{read\_line}(\\{params}\MG\\{dictionary\_file},\39\\{buf})\V\R%
\\{parse\_word}(\\{buf},\39\\{tt},\39\\{params},\39\\{word})\V\R\\{hyphenate%
\_word}(\\{word},\39\\{pt},\39\\{params})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\\{destroy\_word}(\\{word});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{count\_dots}(\\{word},\39\\{params},\39\\{ps});{}$\6
\&{if} ${}(\\{word}\MG\\{length}\G\\{dot\_len}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{process\_word}(\\{word},\39\\{ct},\39\\{params},\39\\{helper%
\_trie})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\\{destroy\_word}(\\{word});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{buf});\6
\\{destroy\_word}(\\{word});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{33}parse\_word.
Parses the text from buffer to {\tt word} structure. The text is translated to
lowercase and hyphens are processed.
Return value indicates successful parsing.

\Y\B\1\1\&{bool} \\{parse\_word}(\&{struct} \\{string\_buffer} ${}{*}\\{buf},%
\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} \\{params}
${}{*}\\{params},\39{}$\&{struct} \\{word} ${}{*}\\{out\_word})\2\2{}$\6
${}\{{}$\1\6
\\{reset\_word}(\\{out\_word});\7
\&{char} \\{edge\_of\_word}[\,]${}\K\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\};{}$\6
\&{struct} \\{string\_buffer} ${}{*}\\{letter}\K\\{init\_buffer}(\T{4});{}$\7
\&{if} ${}(\\{letter}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{uint8\_t}\\{weight}\K\\{params}\MG\\{word\_weight};{}$\7
\&{enum} \\{hyphen\_type} \\{hyf}${}\K\.{NO\_HYF};{}$\6
\&{size\_t} \\{letter\_index}${}\K\\{get\_letter\_index}(\\{tt},\39\\{edge\_of%
\_word});{}$\7
\&{if} ${}(\R\\{convert\_index}(\\{letter\_index},\39\\{out\_word})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{char} \|c;\6
\&{bool} \\{has\_weight}${}\K\\{false};{}$\7
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{buf}\MG\\{size};{}$ ${}\|i%
\PP){}$\5
${}\{{}$\1\6
${}\|c\K\\{buf}\MG\\{data}[\|i];{}$\6
\&{if} (\\{has\_weight})\5
${}\{{}$\1\6
${}\\{weight}\K(\\{uint8\_t})\|c;{}$\6
\&{if} ${}(\|i\E\T{1}){}$\5
${}\{{}$\1\6
${}\\{params}\MG\\{word\_weight}\K\\{weight};{}$\6
\4${}\}{}$\2\6
${}\\{has\_weight}\K\\{false};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\|c\E\.{HYPHEN\_FLAG}){}$\5
${}\{{}$\1\6
${}\\{has\_weight}\K\\{true};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\|c\E\\{params}\MG\\{good\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{hyf}\K\.{GOOD\_HYF};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\|c\E\\{params}\MG\\{missed\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{hyf}\K\.{MISS\_HYF};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\|c\E\\{params}\MG\\{bad\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{hyf}\K\.{BAD\_HYF};{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{is\_utf\_start\_byte}(\|c)\W\\{letter}\MG\\{size}>%
\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\.{'\\0'})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{letter\_index}\K\\{get\_letter\_index}(\\{tt},\39\\{letter}\MG%
\\{data});{}$\6
\&{if} ${}(\\{letter\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Character\ '\%s'\ in\ w}\)\.{ord\ '\%s'\ not\
known\\n}\)\.{"},\39\\{letter}\MG\\{data},\39\\{buf}\MG\\{data});{}$\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{convert\_index}(\\{letter\_index},\39\\{out\_word})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{set\_true\_hyphen}(\\{out\_word},\39\\{out\_word}\MG\\{size}-%
\T{1},\39\T{4}*\\{weight}+\\{hyf})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{hyf}\K\.{NO\_HYF};{}$\6
\\{reset\_buffer}(\\{letter});\6
\4${}\}{}$\2\6
${}\\{weight}\K\\{params}\MG\\{word\_weight};{}$\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\|c)){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{letter}\MG\\{size}>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char}(\\{letter},\39\.{'\\0'})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{letter\_index}\K\\{get\_letter\_index}(\\{tt},\39\\{letter}\MG%
\\{data});{}$\6
\&{if} ${}(\\{letter\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Character\ '\%s'\ in\ w}\)\.{ord\ '\%s'\ not\
known\\n}\)\.{"},\39\\{letter}\MG\\{data},\39\\{buf}\MG\\{data});{}$\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{convert\_index}(\\{letter\_index},\39\\{out\_word})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{letter\_index}\K\\{get\_letter\_index}(\\{tt},\39\\{edge\_of\_word});{}$\6
\&{if} ${}(\R\\{convert\_index}(\\{letter\_index},\39\\{out\_word})\V\R\\{set%
\_true\_hyphen}(\\{out\_word},\39\T{0},\39\T{0})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\\{destroy\_buffer}(\\{letter});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{34}count\_dots.
Collects statistics about the hyphens in parsed word.

\Y\B\1\1\&{void} \\{count\_dots}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{struct} \\{params} ${}{*}\\{params},\39{}$\&{struct} \\{pass\_stats} ${}{*}%
\\{ps})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{word}\MG\\{length}<(\\{uint8\_t})(\\{params}\MG\\{right\_hyphen%
\_min}+\T{1})){}$\5
${}\{{}$\1\6
\&{return};\6
\4${}\}{}$\2\7
\&{size\_t} \\{current\_index}${}\K\T{0};{}$\6
\&{size\_t} \\{current\_pos}${}\K\T{0};{}$\6
\&{bool} \\{odd\_level};\6
\&{size\_t} \\{dot\_index}${},\39\\{hyphenation\_value},\39\\{weight};{}$\6
\&{enum} \\{hyphen\_type} \\{hyf};\7
\&{for} (\&{size\_t} \\{dot\_pos}${}\K\\{params}\MG\\{left\_hyphen\_min}+%
\T{1};{}$ ${}\\{dot\_pos}<(\\{uint8\_t})(\\{word}\MG\\{length}-\\{params}\MG%
\\{right\_hyphen\_min});{}$ ${}\\{dot\_pos}\PP){}$\5
${}\{{}$\1\6
\&{while} ${}(\\{current\_pos}<\\{dot\_pos}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{current\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{current\_index}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{dot\_index}\K\\{current\_index}-\T{1};{}$\6
${}\\{odd\_level}\K(\\{get\_found\_hyphen}(\\{word},\39\\{dot\_index})\MOD\T{2}%
\E\T{1});{}$\6
${}\\{hyphenation\_value}\K\\{get\_true\_hyphen}(\\{word},\39\\{dot%
\_index});{}$\6
${}\\{weight}\K\\{hyphenation\_value}/\T{4};{}$\6
${}\\{hyf}\K\\{hyphenation\_value}\MOD\T{4};{}$\6
\&{if} ${}(\\{hyphenation\_value}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Code\ I\ hoped\ unreac}\)\.{hable\ was\
reached\\n"});{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{hyf}\MOD\T{2}\E\T{0}){}$\5
${}\{{}$\1\6
\&{if} (\\{odd\_level})\5
${}\{{}$\1\6
${}\\{ps}\MG\\{bad\_cnt}\MRL{+{\K}}\\{weight};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{if} (\\{odd\_level})\5
${}\{{}$\1\6
${}\\{ps}\MG\\{good\_cnt}\MRL{+{\K}}\\{weight};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{ps}\MG\\{miss\_cnt}\MRL{+{\K}}\\{weight};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\fi

\M{35}process\_word.
Generates all candidate patterns from the parsed word and inserts them to count
trie. Return value indicates the success
of generating and insertion.

\Y\B\1\1\&{bool} \\{process\_word}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{struct} \\{count\_trie} ${}{*}\\{ct},\39{}$\&{struct} \\{params} ${}{*}%
\\{params},\39{}$\&{struct} \\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
${}\\{uint8\_t}\\{dot\_min}\K\\{params}\MG\\{pat\_dot};{}$\6
${}\\{uint8\_t}\\{dot\_max}\K\\{params}\MG\\{pat\_len}-\\{params}\MG\\{pat%
\_dot};{}$\6
\&{if} ${}(\\{dot\_min}<\\{params}\MG\\{left\_hyphen\_min}+\T{1}){}$\5
${}\{{}$\1\6
${}\\{dot\_min}\K\\{params}\MG\\{left\_hyphen\_min}+\T{1};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{dot\_max}<\\{params}\MG\\{right\_hyphen\_min}+\T{1}){}$\5
${}\{{}$\1\6
${}\\{dot\_max}\K\\{params}\MG\\{right\_hyphen\_min}+\T{1};{}$\6
\4${}\}{}$\2\7
\&{size\_t} \\{start\_pos}${},\39\\{end\_pos},\39\\{start\_index},\39\\{dot%
\_index},\39\\{end\_index},\39\\{node},\39\\{weight},\39\\{cnt\_index};{}$\6
\&{size\_t} \\{current\_pos}${}\K\T{0};{}$\6
\&{size\_t} \\{current\_index}${}\K\T{0};{}$\6
\&{bool} \\{good\_pattern};\6
\&{enum} \\{hyphen\_type} \\{hyf};\7
\&{for} (\&{size\_t} \\{dot\_pos}${}\K\\{dot\_min};{}$ ${}\\{dot\_pos}+\\{dot%
\_max}\Z\\{word}\MG\\{length};{}$ ${}\\{dot\_pos}\PP){}$\5
${}\{{}$\1\6
\&{while} ${}(\\{current\_pos}<\\{dot\_pos}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{current\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{current\_index}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{dot\_index}\K\\{current\_index}-\T{1};{}$\6
\&{if} ${}(\\{get\_no\_more}(\\{word},\39\\{dot\_index})){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{hyf}\K\\{get\_true\_hyphen}(\\{word},\39\\{dot\_index})\MOD\T{4};{}$\6
\&{if} ${}(\\{get\_found\_hyphen}(\\{word},\39\\{dot\_index})\MOD\T{2}\E%
\T{1}){}$\5
${}\{{}$\1\6
${}\\{hyf}\MRL{+{\K}}\T{2};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{hyf}\E\\{params}\MG\\{good\_dot}){}$\5
${}\{{}$\1\6
${}\\{good\_pattern}\K\\{true};{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{hyf}\E\\{params}\MG\\{bad\_dot}){}$\5
${}\{{}$\1\6
${}\\{good\_pattern}\K\\{false};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{start\_pos}\K\\{dot\_pos};{}$\6
${}\\{start\_index}\K\\{current\_index};{}$\6
\&{while} ${}(\\{start\_pos}+\\{params}\MG\\{pat\_dot}\G\\{dot\_pos}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{start\_index}\E\T{0}\V(\\{uint8\_t})\\{word}\MG\\{translated}[%
\\{start\_index}-\T{1}]\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{start\_pos}\MM;{}$\6
\4${}\}{}$\2\6
${}\\{start\_index}\MM;{}$\6
\4${}\}{}$\2\6
${}\\{start\_pos}\PP;{}$\6
${}\\{start\_index}\PP;{}$\6
${}\\{end\_pos}\K\\{dot\_pos};{}$\6
${}\\{end\_index}\K\\{current\_index};{}$\6
\&{while} ${}(\\{end\_index}<\\{word}\MG\\{size}\W\\{end\_pos}<\\{start\_pos}+%
\\{params}\MG\\{pat\_len}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{end\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{end\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{end\_index}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{insert\_substring}(\\{ct}\MG\|t,\39\\{word}\MG\\{translated},%
\39\\{end\_index},\39\\{end\_index}-\\{start\_index},\39{\AND}\\{node},\39%
\\{helper\_trie})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{ct}\MG\\{cnts}\MG\\{size}\G\\{ct}\MG\\{cnts}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\T{2}*\\{ct}\MG\\{cnts}\MG\\{capacity};{}$\7
\&{if} ${}(\\{resize\_pattern\_counts}(\\{ct}\MG\\{cnts},\39\\{new\_capacity})%
\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{cnt\_index}\K\\{ct}\MG\|t\MG\\{aux}[\\{node}];{}$\6
\&{if} ${}(\\{cnt\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{cnt\_index}\K\\{ct}\MG\\{cnts}\MG\\{size};{}$\6
${}\\{ct}\MG\|t\MG\\{aux}[\\{node}]\K\\{cnt\_index};{}$\6
${}\\{ct}\MG\\{cnts}\MG\\{size}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{weight}\K\\{get\_true\_hyphen}(\\{word},\39\\{dot\_index})/\T{4};{}$\6
\&{if} ${}(\\{weight}\E\T{0}){}$\1\5
${}\\{weight}\K\\{params}\MG\\{word\_weight};{}$\2\6
\&{if} (\\{good\_pattern})\5
${}\{{}$\1\6
${}\\{ct}\MG\\{cnts}\MG\\{good}[\\{cnt\_index}]\MRL{+{\K}}\\{weight};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{ct}\MG\\{cnts}\MG\\{bad}[\\{cnt\_index}]\MRL{+{\K}}\\{weight};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{36}Pattern collection.
After the dictionary has been processed, the count trie contains the number of
good (supporting) and bad
(contradicting) occurences of each candidate pattern found. There are 3 types
of patterns based on these counts and the
\goodwtpar{}, \badwtpar{}, and \threshpar{} parameters:

\item{$\bullet$} {\bf good patterns} for which inequality
$
good * good\_wt - bad * bad\_wt \geq thresh
$
holds. Good patterns are inserted into the pattern trie with corresponding
hyphenation level.
\item{$\bullet$} {\bf bad patterns}: for which inequality
$
good * good\_wt < thresh
$
holds. Bad patterns are inserted into the pattern trie with a special {\tt BAD%
\_OP\_VALUE} level of value
255. Neither them nor their superstrings can become good patterns and are
deleted from the pattern trie in the
end of the hyphenation level iteration.
\item{$\bullet$} {\bf undecided patterns}: for which none of the inequalities
above holds. Some of their superstrings may
become good or bad patterns later, so need for further investigation is raised
by setting the
{\tt more\_to\_come} flag.

\fi

\M{37}collect\_count\_trie.
Generates new patterns from the count trie and prints out statistics. Return
value indicates whether the generation
ended successfully.

\Y\B\1\1\&{bool} \\{collect\_count\_trie}(\&{struct} \\{count\_trie} ${}{*}%
\\{ct},\39{}$\&{struct} \\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} %
\\{params} ${}{*}\\{params},\39{}$\&{struct} \\{pass\_stats} ${}{*}\\{ps})\2%
\2{}$\6
${}\{{}$\1\6
\&{double} \\{bad\_eff}${}\K(\&{double})\,\\{params}\MG\\{thresh}/(\&{double})%
\,\\{params}\MG\\{good\_wt};{}$\7
${}\\{ps}\MG\\{good\_pat\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{bad\_pat\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{good\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{bad\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{more\_to\_come}\K\\{false};{}$\6
\&{if} ${}(\R\\{traverse\_count\_trie}(\\{ct},\39\\{pt},\39\\{params},\39%
\\{ps})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\%zu\ good\ and\ \%zu\ ba}\)\.{d\ patterns\ added"},\39\\{ps}%
\MG\\{good\_pat\_cnt},\39\\{ps}\MG\\{bad\_pat\_cnt});{}$\6
${}\\{ps}\MG\\{level\_pattern\_cnt}\MRL{+{\K}}\\{ps}\MG\\{good\_pat\_cnt};{}$\6
\&{if} ${}(\\{ps}\MG\\{more\_to\_come}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"\ (more\ to\ come)\\n"});\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"finding\ \%zu\ good\ an}\)\.{d\ \%zu\ bad\ hyphens"},\39%
\\{ps}\MG\\{good\_cnt},\39\\{ps}\MG\\{bad\_cnt});{}$\6
\&{if} ${}(\\{ps}\MG\\{good\_pat\_cnt}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{",\ efficiency\ =\ \%.2l}\)\.{f\\n"},\39(\&{double})\,\\{ps}%
\MG\\{good\_cnt}/(\\{ps}\MG\\{good\_pat\_cnt}+((\&{double})\,\\{ps}\MG\\{bad%
\_cnt}/\\{bad\_eff})));{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\\{printf}(\.{"\\n"});\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"pattern\ trie\ has\ \%z}\)\.{u\ nodes,\ trie\_max\ =\ }\)\.{%
\%zu,\ \%zu\ outputs\\n"},\39\\{pt}\MG\|t\MG\\{occupied},\39\\{pt}\MG\|t\MG%
\\{node\_max},\39\\{pt}\MG\\{ops}\MG\\{count});{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{38}traverse\_count\_trie.
Searches through the count trie and inserts accepted patterns to the pattern
trie. Statistics are collected along the
way. Return value indicates whether the whole trie was explored.

\Y\B\1\1\&{bool} \\{traverse\_count\_trie}(\&{struct} \\{count\_trie} ${}{*}%
\\{ct},\39{}$\&{struct} \\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} %
\\{params} ${}{*}\\{params},\39{}$\&{struct} \\{pass\_stats} ${}{*}\\{ps})\2%
\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{root}${}\K\T{1};{}$\6
\&{size\_t} \\{current\_len}${}\K\T{0};{}$\7
\\{uint8\_t}\|c;\7
\&{struct} \\{string\_buffer} ${}{*}\\{pattern}\K\\{init\_buffer}(\T{4}*%
\\{params}\MG\\{pat\_len});{}$\7
\&{if} ${}(\\{pattern}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{stack} ${}{*}\\{s\_base}\K\\{init\_stack}(\T{4}*\\{params}\MG%
\\{pat\_len});{}$\7
\&{if} ${}(\\{s\_base}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{pattern});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{trie} ${}{*}\\{helper\_trie}\K\\{init\_trie}(\T{256});{}$\7
\&{if} ${}(\\{helper\_trie}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char}(\\{pattern},\39\.{'\\0'})\V\R\\{put\_on\_stack}(%
\\{s\_base},\39\\{root})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{node}${},\39\\{op\_index},\39\\{good},\39\\{bad},\39\\{cnt%
\_index};{}$\7
\&{while} ${}(\\{s\_base}\MG\\{top}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{root}\K\\{get\_top\_value}(\\{s\_base});{}$\6
${}\\{pattern}\MG\\{data}[\\{pattern}\MG\\{size}-\T{1}]\MRL{+{\K}}\T{1};{}$\6
${}\|c\K(\\{uint8\_t})\\{pattern}\MG\\{data}[\\{pattern}\MG\\{size}-\T{1}];{}$\6
\&{if} ${}(\|c\E\T{0}){}$\5
${}\{{}$\SHC{ overflow }\1\6
${}\\{pattern}\MG\\{size}\MM;{}$\6
${}\\{s\_base}\MG\\{top}\MM;{}$\6
\&{if} ${}(\\{pattern}\MG\\{size}<\T{1}\V(\\{uint8\_t})\\{pattern}\MG\\{data}[%
\\{pattern}\MG\\{size}-\T{1}]\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_len}\MM;{}$\6
\4${}\}{}$\2\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{node}\K\\{root}+\|c;{}$\6
\&{if} ${}((\\{uint8\_t})\\{ct}\MG\|t\MG\\{nodes}[\\{node}]\I\|c){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}((\\{uint8\_t})\|c\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_len}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{current\_len}\E\\{params}\MG\\{pat\_len}){}$\5
${}\{{}$\1\6
${}\\{cnt\_index}\K\\{ct}\MG\|t\MG\\{aux}[\\{node}];{}$\6
${}\\{good}\K\\{ct}\MG\\{cnts}\MG\\{good}[\\{cnt\_index}];{}$\6
${}\\{bad}\K\\{ct}\MG\\{cnts}\MG\\{bad}[\\{cnt\_index}];{}$\6
\&{if} ${}(\\{good}>\T{0}\V\\{bad}>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{params}\MG\\{good\_wt}*\\{good}<\\{params}\MG\\{thresh}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{insert\_substring}(\\{pt}\MG\|t,\39\\{pattern}\MG\\{data},\39%
\\{pattern}\MG\\{size},\39\\{pattern}\MG\\{size},\39{\AND}\\{op\_index},\39%
\\{helper\_trie})\V\R\\{set\_output}(\\{pt},\39\\{op\_index},\39\.{BAD\_OP%
\_VALUE},\39\\{params}\MG\\{pat\_dot})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{ps}\MG\\{bad\_pat\_cnt}\PP;{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{params}\MG\\{good\_wt}*\\{good}\G\\{params}\MG%
\\{thresh}+\\{params}\MG\\{bad\_wt}*\\{bad}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{insert\_substring}(\\{pt}\MG\|t,\39\\{pattern}\MG\\{data},\39%
\\{pattern}\MG\\{size},\39\\{pattern}\MG\\{size},\39{\AND}\\{op\_index},\39%
\\{helper\_trie})\V\R\\{set\_output}(\\{pt},\39\\{op\_index},\39\\{params}\MG%
\\{hyph\_level},\39\\{params}\MG\\{pat\_dot})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{ps}\MG\\{good\_pat\_cnt}\PP;{}$\6
${}\\{ps}\MG\\{good\_cnt}\MRL{+{\K}}\\{good};{}$\6
${}\\{ps}\MG\\{bad\_cnt}\MRL{+{\K}}\\{bad};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{ps}\MG\\{more\_to\_come}\K\\{true};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}((\\{uint8\_t})\|c\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_len}\MM;{}$\6
\4${}\}{}$\2\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{root}\K\\{ct}\MG\|t\MG\\{links}[\\{node}];{}$\6
\&{if} ${}(\\{root}\E\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\|c\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_len}\MM;{}$\6
\4${}\}{}$\2\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char}(\\{pattern},\39\.{'\\0'})\V\R\\{put\_on\_stack}(%
\\{s\_base},\39\\{root})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_trie}(\\{helper\_trie});\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{39}Pattern pruning.
Once all the required pattern lengths and dot positions were explored for given
hyphenation level, the pattern trie is
pruned of the patterns that were marked as bad.

\fi

\M{40}delete\_bad\_patterns.
Removes bad patterns from the pattern trie and prints out statistics. Return
value indicates success of the operation.

\Y\B\1\1\&{bool} \\{delete\_bad\_patterns}(\&{struct} \\{pattern\_trie} ${}{*}%
\\{pt})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{old\_op\_cnt}${}\K\\{pt}\MG\\{ops}\MG\\{count};{}$\6
\&{size\_t} \\{old\_trie\_cnt}${}\K\\{pt}\MG\|t\MG\\{occupied};{}$\7
\&{if} ${}(\R\\{delete\_patterns}(\\{pt})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|h${}\K\T{1};{}$ ${}\|h\Z\\{pt}\MG\\{ops}\MG%
\\{capacity};{}$ ${}\|h\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{data}[\|h].\\{value}\E\.{BAD\_OP\_VALUE}){}$\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\|h].\\{value}\K\.{EMPTY\_OP\_VALUE};{}$\6
${}\\{pt}\MG\\{ops}\MG\\{count}\MM;{}$\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\|h].\\{next\_op\_index}\K\\{pt}\MG\\{ops}\MG%
\\{data}[\T{0}].\\{next\_op\_index};{}$\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\|h;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\%zu\ nodes\ and\ \%zu\ o}\)\.{utputs\ deleted\\n"},\39\\{old%
\_trie\_cnt}-\\{pt}\MG\|t\MG\\{occupied},\39\\{old\_op\_cnt}-\\{pt}\MG\\{ops}%
\MG\\{count});{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{41}delete\_patterns.
Searches through the pattern trie and removes outputs pointing to the bad
patterns. Nodes that are not used after the
deletion are deleted as well. Return value indicates whether the whole trie was
explored.

\Y\B\1\1\&{bool} \\{delete\_patterns}(\&{struct} \\{pattern\_trie} ${}{*}%
\\{pt})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{root}${}\K\T{1};{}$\6
\&{struct} \\{stack} ${}{*}\\{s\_base}\K\\{init\_stack}(\T{16});{}$\7
\&{if} ${}(\\{s\_base}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{stack} ${}{*}\\{s\_offset}\K\\{init\_stack}(\T{16});{}$\7
\&{if} ${}(\\{s\_offset}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{stack} ${}{*}\\{s\_freed}\K\\{init\_stack}(\T{16});{}$\7
\&{if} ${}(\\{s\_freed}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{put\_on\_stack}(\\{s\_base},\39\\{root})\V\R\\{put\_on\_stack}(%
\\{s\_offset},\39\T{0})\V\R\\{put\_on\_stack}(\\{s\_freed},\39(\&{size\_t})\,%
\\{true})){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\\{destroy\_stack}(\\{s\_freed});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{node};\7
\\{uint8\_t}\|c;\6
\&{while} ${}(\\{s\_base}\MG\\{top}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{root}\K\\{get\_top\_value}(\\{s\_base});{}$\6
${}\\{set\_top\_value}(\\{s\_offset},\39(\\{uint8\_t})\\{get\_top\_value}(\\{s%
\_offset})+\T{1});{}$\6
${}\|c\K(\\{uint8\_t})\\{get\_top\_value}(\\{s\_offset});{}$\6
\&{if} ${}(\|c\E\T{0}){}$\5
${}\{{}$\1\6
\&{bool} \\{child\_freed}${}\K(\\{get\_top\_value}(\\{s\_freed})\E(\&{size\_t})%
\,\\{true});{}$\7
\&{if} (\\{child\_freed})\5
${}\{{}$\1\6
\&{if} ${}(\R\\{set\_base\_used}(\\{pt}\MG\|t,\39\\{root},\39\\{false})){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\\{destroy\_stack}(\\{s\_freed});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{s\_offset}\MG\\{top}\MM;{}$\6
${}\\{s\_base}\MG\\{top}\MM;{}$\6
${}\\{s\_freed}\MG\\{top}\MM;{}$\6
\&{if} ${}(\\{s\_base}\MG\\{top}>\T{0}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{parent\_root}${}\K\\{get\_top\_value}(\\{s\_base});{}$\7
${}\\{uint8\_t}\\{parent\_c}\K(\\{uint8\_t})\\{get\_top\_value}(\\{s%
\_offset});{}$\7
\&{size\_t} \\{parent\_node}${}\K\\{parent\_root}+\\{parent\_c};{}$\7
\&{if} (\\{child\_freed})\5
${}\{{}$\1\6
${}\\{pt}\MG\|t\MG\\{links}[\\{parent\_node}]\K\T{0};{}$\6
\&{if} ${}(\\{pt}\MG\|t\MG\\{aux}[\\{parent\_node}]\E\T{0}\W\\{parent\_root}\I%
\T{1}){}$\5
${}\{{}$\1\6
${}\\{deallocate\_node}(\\{pt}\MG\|t,\39\\{parent\_node});{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{set\_top\_value}(\\{s\_freed},\39(\&{size\_t})\,\\{false});{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{set\_top\_value}(\\{s\_freed},\39(\&{size\_t})\,\\{false});{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{node}\K\\{root}+\|c;{}$\6
\&{if} ${}((\\{uint8\_t})\\{pt}\MG\|t\MG\\{nodes}[\\{node}]\I\|c){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{link\_around\_bad\_outputs}(\\{pt},\39\\{node})){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\\{destroy\_stack}(\\{s\_freed});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{pt}\MG\|t\MG\\{aux}[\\{node}]>\T{0}\V\\{root}\E\T{1}){}$\5
${}\{{}$\1\6
${}\\{set\_top\_value}(\\{s\_freed},\39(\&{size\_t})\,\\{false});{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{if} ${}(\\{pt}\MG\|t\MG\\{links}[\\{node}]\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{deallocate\_node}(\\{pt}\MG\|t,\39\\{node});{}$\6
\&{continue};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{root}\K\\{pt}\MG\|t\MG\\{links}[\\{node}];{}$\6
\&{if} ${}(\\{root}\E\T{0}){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{put\_on\_stack}(\\{s\_base},\39\\{root})\V\R\\{put\_on\_stack}(%
\\{s\_offset},\39\T{0})\V\R\\{put\_on\_stack}(\\{s\_freed},\39(\&{size\_t})\,%
\\{true})){}$\5
${}\{{}$\1\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\\{destroy\_stack}(\\{s\_freed});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_stack}(\\{s\_base});\6
\\{destroy\_stack}(\\{s\_offset});\6
\\{destroy\_stack}(\\{s\_freed});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{42}deallocate\_node.
Removes unused node from a trie and links it to the beginning of free space
chain.

\Y\B\1\1\&{void} \\{deallocate\_node}(\&{struct} \\{trie} ${}{*}\|t,\39{}$%
\&{size\_t} \\{t\_index})\2\2\6
${}\{{}$\1\6
\&{size\_t} \\{old\_head}${}\K\|t\MG\\{links}[\T{0}];{}$\7
${}\|t\MG\\{nodes}[\\{t\_index}]\K\T{0};{}$\6
${}\|t\MG\\{links}[\T{0}]\K\\{t\_index};{}$\6
${}\|t\MG\\{aux}[\\{t\_index}]\K\T{0};{}$\6
${}\|t\MG\\{links}[\\{t\_index}]\K\\{old\_head};{}$\6
${}\|t\MG\\{aux}[\\{old\_head}]\K\\{t\_index};{}$\6
${}\|t\MG\\{occupied}\MM;{}$\6
\4${}\}{}$\2\par
\fi

\M{43}link\_around\_bad\_outputs.
Fixes the output linking to not contain the outputs that will be deleted.
Returns true upon success.

\Y\B\1\1\&{bool} \\{link\_around\_bad\_outputs}(\&{struct} \\{pattern\_trie}
${}{*}\\{pt},\39{}$\&{size\_t} \\{t\_index})\2\2\6
${}\{{}$\1\6
\&{size\_t} \\{lookup\_index}${}\K\\{pt}\MG\|t\MG\\{aux}[\\{t\_index}];{}$\7
\&{if} ${}(\\{lookup\_index}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\&{size\_t} \\{op\_index}${}\K\\{pt}\MG\\{ops}\MG\\{lookup}[\\{lookup%
\_index}];{}$\6
\&{size\_t} \\{free\_list\_head}${}\K\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next%
\_op\_index};{}$\6
\&{size\_t} \|h${}\K\T{0};{}$\7
${}\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\\{op\_index};{}$\7
\&{size\_t} \|n${}\K\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index};{}$\7
\&{while} ${}(\|n>\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{data}[\|n].\\{value}\E\.{BAD\_OP\_VALUE}){}$\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\|h].\\{next\_op\_index}\K\\{pt}\MG\\{ops}\MG%
\\{data}[\|n].\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\|h\K\|n;{}$\6
\4${}\}{}$\2\6
${}\|n\K\\{pt}\MG\\{ops}\MG\\{data}[\|h].\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|h\E\T{0}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{lookup}[\\{lookup\_index}]>\T{0}){}$\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{lookup}[\\{lookup\_index}]\K\T{0};{}$\6
${}\\{pt}\MG\\{ops}\MG\\{lookup\_cnt}\MM;{}$\6
\4${}\}{}$\2\6
${}\\{pt}\MG\|t\MG\\{aux}[\\{t\_index}]\K\T{0};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{lookup}[\\{lookup\_index}]\K\\{pt}\MG\\{ops}\MG%
\\{data}[\T{0}].\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\\{free\_list%
\_head};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{44}Pattern output.
At the end of the run, \utfpatgen{} prints out the final set of patterns to the
output file. The format is the same as
the required format of the input pattern file -- each value of a hyphenation
level is preceded by the
{\tt HYPHEN\_FLAG} symbol, the '.' character marking an edge of word.

\fi

\M{45}output\_patterns.
Traverses through the pattern trie and writes all the patterns present there to
the output file. Returns true if the
whole trie has been explored and patterns successfully written.

\Y\B\1\1\&{bool} \\{output\_patterns}(\&{struct} \\{pattern\_trie} ${}{*}%
\\{pt},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{FILE} ${}{*}%
\\{output\_file})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{root}${}\K\T{1};{}$\7
\\{uint8\_t}\|c;\7
\&{struct} \\{string\_buffer} ${}{*}\\{pattern}\K\\{init\_buffer}(\T{16});{}$\7
\&{if} ${}(\\{pattern}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{stack} ${}{*}\\{s\_base}\K\\{init\_stack}(\T{16});{}$\7
\&{if} ${}(\\{s\_base}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{pattern});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char}(\\{pattern},\39\.{'\\0'})\V\R\\{put\_on\_stack}(%
\\{s\_base},\39\\{root})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{node};\7
\&{while} ${}(\\{s\_base}\MG\\{top}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{root}\K\\{get\_top\_value}(\\{s\_base});{}$\6
${}\\{pattern}\MG\\{data}[\\{pattern}\MG\\{size}-\T{1}]\MRL{+{\K}}\T{1};{}$\6
${}\|c\K(\\{uint8\_t})\\{pattern}\MG\\{data}[\\{pattern}\MG\\{size}-\T{1}];{}$\6
\&{if} ${}(\|c\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{pattern}\MG\\{data}[\\{pattern}\MG\\{size}-\T{1}]\K\.{'\\0'};{}$\6
${}\\{pattern}\MG\\{size}\MM;{}$\6
${}\\{s\_base}\MG\\{top}\MM;{}$\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{node}\K\\{root}+\|c;{}$\6
\&{if} ${}((\\{uint8\_t})\\{pt}\MG\|t\MG\\{nodes}[\\{node}]\I\|c){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{pt}\MG\|t\MG\\{aux}[\\{node}]>\T{0}){}$\5
${}\{{}$\1\6
${}\\{output\_pattern}(\\{pattern},\39\\{tt},\39\\{pt}\MG\\{ops},\39\\{pt}\MG%
\|t\MG\\{aux}[\\{node}],\39\\{output\_file});{}$\6
\4${}\}{}$\2\6
${}\\{root}\K\\{pt}\MG\|t\MG\\{links}[\\{node}];{}$\6
\&{if} ${}(\\{root}\E\T{0}){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char}(\\{pattern},\39\.{'\\0'})\V\R\\{put\_on\_stack}(%
\\{s\_base},\39\\{root})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_buffer}(\\{pattern});\6
\\{destroy\_stack}(\\{s\_base});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{46}output\_pattern.
Writes single pattern to the output trie.

\Y\B\1\1\&{void} \\{output\_pattern}(\&{struct} \\{string\_buffer} ${}{*}%
\\{pattern},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct}
\\{outputs} ${}{*}\\{ops},\39{}$\&{size\_t} \\{op\_index}${},\39{}$\&{FILE}
${}{*}\\{output\_file})$ $\{$ \6
\&{if} ${}(\\{op\_index}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return};\6
\4${}\}{}$\2\7
\&{size\_t} \\{pattern\_position}${}\K\T{0};{}$\6
\&{size\_t} \\{level}${},\39\\{letter\_index};{}$\6
\&{char} ${}{*}\\{word\_index}\K\\{pattern}\MG\\{data};$ \&{size\_t} \\{edge%
\_of\_word\_idx} $\K$ \\{get\_letter\_index} $(\\{tt},\39$ \1\1(\&{char}[%
\T{2}]) $\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\}$ )  ;\6
\&{while} ${}(\\{word\_index}<\\{pattern}\MG\\{data}+\\{pattern}\MG\\{size}){}$%
\5
${}\{{}$\1\6
${}\\{level}\K\\{get\_highest\_level}(\\{ops},\39\\{op\_index},\39\\{pattern%
\_position});{}$\6
\&{if} ${}(\\{level}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{fputc}(\.{'\\xfe'},\39\\{output\_file});{}$\6
${}\\{fputc}((\\{uint8\_t})\\{level},\39\\{output\_file});{}$\6
\4${}\}{}$\2\6
${}\\{letter\_index}\K\\{convert\_byte\_sequence}({\AND}\\{word\_index});{}$\6
\&{if} ${}(\\{letter\_index}\E\\{edge\_of\_word\_idx}){}$\5
${}\{{}$\1\6
${}\\{fputc}(\.{'.'},\39\\{output\_file});{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{fprintf}(\\{output\_file},\39\.{"\%s"},\39\\{tt}\MG\\{alphabet}\MG%
\\{data}+\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]);{}$\6
\4${}\}{}$\2\6
${}\\{pattern\_position}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{level}\K\\{get\_highest\_level}(\\{ops},\39\\{op\_index},\39\\{pattern%
\_position});{}$\6
\&{if} ${}(\\{level}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{fputc}(\.{'\\xfe'},\39\\{output\_file});{}$\6
${}\\{fputc}((\\{uint8\_t})\\{level},\39\\{output\_file});{}$\6
\4${}\}{}$\2\6
${}\\{fputc}(\.{'\\n'},\39\\{output\_file});$ $\}{}$\par
\fi

\M{47}get\_highest\_level.
Selects the hyphenation level to be used in given dot position. Return the
level value.

\Y\B\1\1\&{size\_t} \\{get\_highest\_level}(\&{struct} \\{outputs} ${}{*}%
\\{ops},\39{}$\&{size\_t} \\{start\_index}${},\39{}$\&{size\_t} \\{position})\2%
\2\6
${}\{{}$\1\6
\&{size\_t} \\{highest}${}\K\T{0};{}$\6
\&{size\_t} \\{op\_index}${}\K\\{ops}\MG\\{lookup}[\\{start\_index}];{}$\6
\&{struct} \\{output} \\{op};\7
\&{while} ${}(\\{op\_index}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{op}\K\\{ops}\MG\\{data}[\\{op\_index}];{}$\6
\&{if} ${}(\\{op}.\\{position}\E\\{position}\W\\{op}.\\{value}\I\.{BAD\_OP%
\_VALUE}\W\\{op}.\\{value}>\\{highest}){}$\5
${}\{{}$\1\6
${}\\{highest}\K\\{op}.\\{value};{}$\6
\4${}\}{}$\2\6
${}\\{op\_index}\K\\{op}.\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
\&{return} \\{highest};\6
\4${}\}{}$\2\par
\fi

\N{1}{48}Hyphenation.
Hyphenation appears on several places in the algorithm. Already in the
dictionary processing, the words are hyphenated
with the current set of patterns. The other case occurs at the absolute end of %
\utfpatgen{} when the user wishes to see
the hyphenated dictionary -- the final set of patterns is used to hyphenate the
dictionary entries into 'pattmp.X'
file. The 'X' marks the last hyphenation level explored.

\fi

\M{49}hyphenate\_word.
Hyphenates the given word according to the actual set of patterns from pattern
trie. Found hyphens are stored in word's
{\tt found\_hyphens} field. Returns true if hyphenation ended without errors.

\Y\B\1\1\&{bool} \\{hyphenate\_word}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{struct} \\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{struct} \\{params} ${}{*}%
\\{params})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{current\_index}${}\K\T{0};{}$\6
\&{size\_t} \\{current\_pos}${}\K\T{0};{}$\6
\&{size\_t} \\{node}${},\39\\{base},\39\\{start\_index},\39\\{dot\_index},\39%
\\{end\_index},\39\\{op\_index},\39\\{dot\_pos},\39\\{end\_pos};{}$\6
\&{struct} \\{output} \\{op};\7
\&{if} ${}(\\{word}\MG\\{length}<(\\{uint8\_t})(\\{params}\MG\\{right\_hyphen%
\_min}+\T{1})){}$\5
${}\{{}$\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\&{size\_t} \\{start\_pos}${}\K\T{0};{}$\7
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{word}\MG\\{length}-\\{params}%
\MG\\{right\_hyphen\_min};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
\&{while} ${}(\\{current\_pos}<\\{start\_pos}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{current\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{current\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{current\_index}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{start\_index}\K\\{current\_index};{}$\6
${}\\{end\_index}\K\\{current\_index};{}$\6
${}\\{end\_pos}\K\\{current\_pos};{}$\6
${}\\{node}\K\T{1}+(\\{uint8\_t})\\{word}\MG\\{translated}[\\{start%
\_index}];{}$\6
\&{while} ${}(\\{pt}\MG\|t\MG\\{nodes}[\\{node}]\E\\{word}\MG\\{translated}[%
\\{end\_index}]){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{end\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{end\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{end\_index}\PP;{}$\6
${}\\{op\_index}\K\\{pt}\MG\\{ops}\MG\\{lookup}[\\{pt}\MG\|t\MG\\{aux}[%
\\{node}]];{}$\6
\&{while} ${}(\\{op\_index}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{op}\K\\{pt}\MG\\{ops}\MG\\{data}[\\{op\_index}];{}$\6
${}\\{dot\_pos}\K\\{start\_pos};{}$\6
${}\\{dot\_index}\K\\{start\_index};{}$\6
\&{while} ${}(\\{dot\_pos}<\\{start\_pos}+\\{op}.\\{position}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{uint8\_t})\\{word}\MG\\{translated}[\\{dot\_index}]\I\T{%
\^ff}){}$\5
${}\{{}$\1\6
${}\\{dot\_pos}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{dot\_index}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{dot\_index}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{dot\_index}\MM;{}$\6
\&{if} ${}(\\{op}.\\{value}<\.{BAD\_OP\_VALUE}\W\\{get\_found\_hyphen}(%
\\{word},\39\\{dot\_index})<\\{op}.\\{value}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{set\_found\_hyphen}(\\{word},\39\\{dot\_index},\39\\{op}.%
\\{value})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{op}.\\{value}\G\\{params}\MG\\{hyph\_level}){}$\5
${}\{{}$\1\6
\&{if} ${}((\\{end\_pos}+\\{params}\MG\\{pat\_dot}\Z\\{dot\_pos}+\\{params}\MG%
\\{pat\_len})\W(\\{dot\_pos}\Z\\{start\_pos}+\\{params}\MG\\{pat\_dot})){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{set\_no\_more}(\\{word},\39\\{dot\_index},\39\\{true})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{op\_index}\K\\{op}.\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
${}\\{base}\K\\{pt}\MG\|t\MG\\{links}[\\{node}];{}$\6
\&{if} ${}(\\{base}\E\T{0}){}$\5
${}\{{}$\1\6
\&{break};\6
\4${}\}{}$\2\6
${}\\{node}\K\\{base}+(\\{uint8\_t})\\{word}\MG\\{translated}[\\{end%
\_index}];{}$\6
\4${}\}{}$\2\6
${}\\{start\_pos}\PP;{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{50}hyphenate\_dicitonary.
Hyphenates dictionary entries and writes them to newly created {\tt pattmp}
file if required. Return value indicates
whether the hyphenation and file printout finished successfully.

\Y\B\1\1\&{bool} \\{hyphenate\_dictionary}(\&{struct} \\{params} ${}{*}%
\\{params},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} %
\\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{bool} \\{output}${},\39{}$\&{struct} %
\\{pass\_stats} ${}{*}\\{ps})\2\2{}$\6
${}\{{}$\1\6
${}\\{ps}\MG\\{good\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{bad\_cnt}\K\T{0};{}$\6
${}\\{ps}\MG\\{miss\_cnt}\K\T{0};{}$\6
${}\\{params}\MG\\{word\_weight}\K\T{1};{}$\7
\&{FILE} ${}{*}\\{pattmp}\K\NULL;{}$\7
\&{if} (\\{output})\5
${}\{{}$\1\6
\&{char} ${}{*}\\{filename}\K\\{malloc}(\T{11}*\&{sizeof}(\&{char}));{}$\7
\&{if} ${}(\\{filename}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{sprintf}(\\{filename},\39\.{"pattmp.\%u"},\39\\{params}\MG\\{hyph%
\_level});{}$\6
${}\\{pattmp}\K\\{fopen}(\\{filename},\39\.{"w"});{}$\6
\&{if} ${}(\\{pattmp}\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{filename});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"writing\ \%s\\n"},\39\\{filename});{}$\6
\\{free}(\\{filename});\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{hyphenate\_all\_words}(\\{params},\39\\{tt},\39\\{pt},\39%
\\{pattmp},\39\\{ps})){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{pattmp}\I\NULL){}$\5
${}\{{}$\1\6
\\{fclose}(\\{pattmp});\6
\4${}\}{}$\2\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"\\n\%zu\ good,\ \%zu\ bad}\)\.{,\ \%zu\ missed\\n"},\39\\{ps}%
\MG\\{good\_cnt},\39\\{ps}\MG\\{bad\_cnt},\39\\{ps}\MG\\{miss\_cnt});{}$\6
\&{if} ${}(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG\\{miss\_cnt}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%.2f\ \%\%,\ \%.2f\ \%\%,\ \%}\)\.{.2f\ \%\%\\n"},\39(%
\T{100}*(\&{float})\,\\{ps}\MG\\{good\_cnt}/(\&{float})(\\{ps}\MG\\{good\_cnt}+%
\\{ps}\MG\\{miss\_cnt})),\39(\T{100}*(\&{float})\,\\{ps}\MG\\{bad\_cnt}/(%
\&{float})(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG\\{miss\_cnt})),\39(\T{100}*(%
\&{float})\,\\{ps}\MG\\{miss\_cnt}/(\&{float})(\\{ps}\MG\\{good\_cnt}+\\{ps}\MG%
\\{miss\_cnt})));{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{pattmp}\I\NULL){}$\5
${}\{{}$\1\6
\\{fclose}(\\{pattmp});\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{51}hyphenate\_all\_words.
Itearates over the words in the dictionary, parses, hyphenates, and writes them
to the {\tt pattmp}
file. Returns true if no error occurs.

\Y\B\1\1\&{bool} \\{hyphenate\_all\_words}(\&{struct} \\{params} ${}{*}%
\\{params},\39{}$\&{struct} \\{translate\_table} ${}{*}\\{tt},\39{}$\&{struct} %
\\{pattern\_trie} ${}{*}\\{pt},\39{}$\&{FILE} ${}{*}\\{pattmp},\39{}$\&{struct}
\\{pass\_stats} ${}{*}\\{ps})\2\2{}$\6
${}\{{}$\1\6
${}\\{rewind}(\\{params}\MG\\{dictionary\_file});{}$\7
\&{struct} \\{string\_buffer} ${}{*}\\{buf}\K\\{init\_buffer}(\T{64});{}$\7
\&{if} ${}(\\{buf}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{struct} \\{word} ${}{*}\\{word}\K\\{init\_word}(\T{16});{}$\7
\&{if} ${}(\\{word}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{buf});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{eof}\K\\{false};{}$\6
\&{while} ${}(\R\\{buf}\MG\\{eof}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{read\_line}(\\{params}\MG\\{dictionary\_file},\39\\{buf})\V\R%
\\{parse\_word}(\\{buf},\39\\{tt},\39\\{params},\39\\{word})\V\R\\{hyphenate%
\_word}(\\{word},\39\\{pt},\39\\{params})){}$\5
${}\{{}$\1\6
\\{destroy\_buffer}(\\{buf});\6
\\{destroy\_word}(\\{word});\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{count\_dots}(\\{word},\39\\{params},\39\\{ps});{}$\6
\&{if} ${}(\\{pattmp}\I\NULL\W\\{word}\MG\\{length}>\T{2}){}$\5
${}\{{}$\1\6
${}\\{output\_hyphenated\_word}(\\{pattmp},\39\\{word},\39\\{tt},\39%
\\{params});{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\\{destroy\_buffer}(\\{buf});\6
\\{destroy\_word}(\\{word});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{52}output\_hyphenated\_word.
Writes the parsed and hyphenated word to the {\tt pattmp} file.

\Y\B\1\1\&{void} \\{output\_hyphenated\_word}(\&{FILE} ${}{*}\\{pattmp},\39{}$%
\&{struct} \\{word} ${}{*}\\{word},\39{}$\&{struct} \\{translate\_table} ${}{*}%
\\{tt},\39{}$\&{struct} \\{params} ${}{*}\\{params})$ $\{$ \6
\&{if} ${}(\\{params}\MG\\{word\_weight}>\T{1}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{pattmp},\39\.{"\%d"},\39\\{params}\MG\\{word\_weight});{}$\6
\4${}\}{}$\2\7
\&{size\_t} \\{weight}${},\39\\{letter\_index},\39\\{dot\_pos}\K\T{0};$ \&{size%
\_t} \\{edge\_of\_word\_idx} $\K$ \\{get\_letter\_index} $(\\{tt},\39$ \1\1(%
\&{char}[\T{2}]) $\{\.{EDGE\_OF\_WORD},\39\.{'\\0'}\}$ )  ;\7
\&{char} ${}{*}\\{word\_index}\K\\{word}\MG\\{translated};{}$\6
\&{bool} \\{has\_hyphen}${},\39\\{found\_hyphen};{}$\7
\&{while} ${}(\\{word\_index}<\\{word}\MG\\{translated}+\\{word}\MG\\{size}){}$%
\5
${}\{{}$\1\6
${}\\{found\_hyphen}\K(\\{get\_found\_hyphen}(\\{word},\39\\{word\_index}-%
\\{word}\MG\\{translated})\MOD\T{2}\E\T{1});{}$\6
${}\\{weight}\K\\{get\_true\_hyphen}(\\{word},\39\\{word\_index}-\\{word}\MG%
\\{translated});{}$\6
${}\\{has\_hyphen}\K(\\{weight}\MOD\T{2}\E\T{1});{}$\6
${}\\{letter\_index}\K\\{convert\_byte\_sequence}({\AND}\\{word\_index});{}$\6
\&{if} ${}(\\{letter\_index}\E\\{edge\_of\_word\_idx}){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{fprintf}(\\{pattmp},\39\.{"\%s"},\39\\{tt}\MG\\{alphabet}\MG\\{data}+%
\\{tt}\MG\\{index\_to\_alphabet}[\\{letter\_index}]);{}$\6
${}\\{dot\_pos}\PP;{}$\6
\&{if} ${}(\\{weight}\E\T{0}\V\\{dot\_pos}<\\{params}\MG\\{left\_hyphen\_min}\V%
\\{dot\_pos}\G\\{word}\MG\\{length}-\\{params}\MG\\{right\_hyphen\_min}-%
\T{1}){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{weight}\MRL{{/}{\K}}\T{4};{}$\6
\&{if} ${}(\\{weight}\I\\{params}\MG\\{word\_weight}){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{pattmp},\39\.{"\%zu"},\39\\{weight});{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{found\_hyphen}\W\\{has\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{fputc}((\&{char})\,\\{params}\MG\\{good\_hyphen},\39\\{pattmp});{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{found\_hyphen}\W\R\\{has\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{fputc}((\&{char})\,\\{params}\MG\\{bad\_hyphen},\39\\{pattmp});{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\R\\{found\_hyphen}\W\\{has\_hyphen}){}$\5
${}\{{}$\1\6
${}\\{fputc}((\&{char})\,\\{params}\MG\\{missed\_hyphen},\39\\{pattmp});{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{fputc}(\.{'\\n'},\39\\{pattmp});$ $\}{}$\par
\fi

\N{1}{53}UTF-8 specifics.
Following methods greatly simplify dealing with UTF-8 encoding. We took
advantage of the design that allows to easily
determine whether a byte is the first one in UTF-8 character -- its two highest
bits are not '10'. The first byte also
encodes the number of bytes that form the UTF-8 character. If the highest bit
is '0', the character comprise the single
byte and its meaning is the same as it would be in ASCII encoding. If the
highest bit is '1', the number of '1' bits on
the highest positions equals the number of bytes of the character (with the
exception of '10' which is not allowed in
the leading byte), e.g., the byte of binary value '1110XXXX' is a beginning of
3-byte UTF-8 character. Although bytes
'0xfe' and '0xff' are not formally allowed in UTF-8, we treat them as one-byte
characters with special meaning
({\tt HYPHEN\_FLAG}, {\tt EDGE\_OF\_WORD} respectively).

\fi

\M{54}is\_utf\_start\_byte.
Returns true if the given byte is the start byte of a UTF-8 character or a %
\utfpatgen{} special symbol.

\Y\B\1\1\&{inline} \&{bool} \\{is\_utf\_start\_byte}(\\{uint8\_t}\\{byte})\2\2\6
${}\{{}$\1\6
\&{return} \\{byte}${}<\T{\^80}\V\\{byte}>\T{\^bf};{}$\6
\4${}\}{}$\2\par
\fi

\N{0}{55}Structures.
This section focuses on the data structures we used in \utfpatgen{} and their
components. Here you can find information
about the basic structures such as tries, buffers, and stacks, as well as
composite structures like the pattern and
count trie, translate table, or word.

We tried to name the methods for basic manipulation with the structures
consistently across the whole implementation.
Thanks to that, you will find several methods with common prefixes: {\tt init%
\_XX} performs initial allocation of
resources for structure {\it XX}, {\tt resize\_XX} reallocates them if the
capacity threshold is reached,
{\tt reset\_XX} reverts the structure to initial state, and {\tt destroy\_XX}
frees all the resources to avoid
memory leaks. Furthermore, we use {\tt get\_YY} and {\tt set\_YY} methods to
read and write into the fields of
given structure instead of direct access. This allows us to perform additional
checks on the input values.

\fi

\N{1}{56}Trie.
The packed trie structure forms the backbone of both \patgen{} and \utfpatgen{}
algorithms. The implementation is very
similar in both cases, a set of arrays representing a n-ary tree, condensed for
better space effectiveness. Single node
of a trie comprises a {\it value}, {\it link} and {\it aux} pointers, and {\it
base} indicator. Every admissible index
of the arrays is either occupied by a node, or it is an empty space (its value
is 0). Index 0 is left empty and serves
as the head of empty space chain, index 1 is the root node of the trie itself.

If we want to traverse the trie, that means finding the node corresponding to
given sequence of values $x_1 \dots x_n$
(if such node exists), we start in the root node $r_0$. Our next destination is
node $n_1 = r_0 + x_1$. Then we check
whether $value(n_1) = x_1$ and if the equation holds, we set $r_1 = link(n_1)$
as the new root. If the equation does
not hold or $r_1 = 0$, we can end the search and conclude that the sequence is
not present in the trie. Otherwise, we
repeat the the steps for $n_2 \dots n_n$ and return the $n_n$ node as desired
result.

The internal representation of the trie consists a structure with following
fields:

\item{$\bullet$} {\bf capacity}: maximal node index that can be inserted,
\item{$\bullet$} {\bf occupied}: number of nodes currently in the trie,
\item{$\bullet$} {\bf node\_max}: the highest occupied node index,
\item{$\bullet$} {\bf base\_max}: the highest base index,
\item{$\bullet$} {\bf pattern\_count}: number of patterns (sequences) in the
trie,
\item{$\bullet$} {\bf nodes}: array of values,
\item{$\bullet$} {\bf links}: array of links, pointing either to the base of
occupied node's subtree, or to the
next empty index in case of an empty space,
\item{$\bullet$} {\bf aux}: array of links, pointing either to occupied node's
output, or to the previous empty
index in case of an empty space,
\item{$\bullet$} {\bf taken}: bit array storing the information about whether
indices are used as bases.

\Y\B\1\1\&{struct} \\{trie} ${}{*}{}$\\{init\_trie}(\&{size\_t} \\{capacity})\2%
\2\6
${}\{{}$\1\6
\&{struct} \\{trie} ${}{*}\|t\K{}$\\{malloc}(\&{sizeof}(\&{struct} \\{trie}));\7
\&{if} ${}(\|t\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{capacity}\K\\{capacity};{}$\6
${}\|t\MG\\{nodes}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{char}));{}$\6
${}\|t\MG\\{links}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{size\_t}));{}$\6
${}\|t\MG\\{aux}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{size\_t}));{}$\6
${}\|t\MG\\{taken}\K\\{calloc}((\\{capacity}/\T{8})+\T{1},\39\&{sizeof}(%
\&{char})){}$;\SHC{ bit array }\6
\&{if} ${}(\|t\MG\\{nodes}\E\NULL\V\|t\MG\\{links}\E\NULL\V\|t\MG\\{aux}\E\NULL%
\V\|t\MG\\{taken}\E\NULL\V\R\\{set\_base\_used}(\|t,\39\T{1},\39\\{true})){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
${}\\{free}(\|t\MG\\{nodes});{}$\6
${}\\{free}(\|t\MG\\{links});{}$\6
${}\\{free}(\|t\MG\\{aux});{}$\6
${}\\{free}(\|t\MG\\{taken});{}$\6
\\{free}(\|t);\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{node\_max}\K\T{0};{}$\6
${}\|t\MG\\{base\_max}\K\T{0};{}$\6
${}\|t\MG\\{occupied}\K\T{0};{}$\6
${}\|t\MG\\{pattern\_count}\K\T{0};{}$\6
\\{relink\_trie}(\|t);\6
\&{return} \|t;\6
\4${}\}{}$\2\par
\fi

\M{57}put\_first\_level.
Inserts the set of one-byte sequences '0x01' to '0xff' into the trie. It can be
called during trie initialization,
since these nodes will never be moved elsewhere. Return value indicates whether
all the insertions succeeded.

\Y\B\1\1\&{bool} \\{put\_first\_level}(\&{struct} \\{trie} ${}{*}\|t)\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{root}${}\K\T{1};{}$\6
\&{size\_t} \\{n\_bytes}${}\K\T{255};{}$\7
\&{if} ${}(\|t\MG\\{capacity}<\\{n\_bytes}+\T{2}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K(((\\{n\_bytes}+\T{2})/\|t\MG\\{capacity})+%
\T{1})*\|t\MG\\{capacity};{}$\7
\&{if} ${}(\\{resize\_trie}(\|t,\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{1};{}$ ${}\|i\Z\\{n\_bytes};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
${}\|t\MG\\{nodes}[\\{root}+\|i]\K(\\{uint8\_t})\|i;{}$\6
${}\|t\MG\\{links}[\\{root}+\|i]\K\T{0};{}$\6
${}\|t\MG\\{aux}[\\{root}+\|i]\K\T{0};{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{node\_max}\K\\{root}+\\{n\_bytes};{}$\6
${}\|t\MG\\{base\_max}\K\\{root};{}$\6
${}\|t\MG\\{occupied}\K\\{n\_bytes};{}$\6
\&{if} ${}(\R\\{set\_base\_used}(\|t,\39\\{root},\39\\{true})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{set\_links}(\|t,\39\T{0},\39\|t\MG\\{node\_max}+\T{1});{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{trie} ${}{*}{}$\\{resize\_trie}(\&{struct} \\{trie} ${}{*}%
\|t,\39{}$\&{size\_t} \\{new\_capacity})\2\2\6
${}\{{}$\1\6
\&{void} ${}{*}\\{new\_nodes}\K\\{realloc}(\|t\MG\\{nodes},\39\\{new%
\_capacity}*\&{sizeof}(\&{char}));{}$\7
\&{if} ${}(\\{new\_nodes}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{nodes}\K\\{new\_nodes};{}$\7
\&{size\_t} ${}{*}\\{new\_links}\K\\{realloc}(\|t\MG\\{links},\39\\{new%
\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_links}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{links}\K\\{new\_links};{}$\7
\&{size\_t} ${}{*}\\{new\_aux}\K\\{realloc}(\|t\MG\\{aux},\39\\{new\_capacity}*%
\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_aux}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{aux}\K\\{new\_aux};{}$\7
\&{size\_t} \\{old\_taken\_bytes}${}\K(\|t\MG\\{capacity}/\T{8})+\T{1};{}$\6
\&{size\_t} \\{new\_taken\_bytes}${}\K(\\{new\_capacity}/\T{8})+\T{1};{}$\6
\&{char} ${}{*}\\{new\_taken}\K\\{realloc}(\|t\MG\\{taken},\39\\{new\_taken%
\_bytes}*\&{sizeof}(\&{char}));{}$\7
\&{if} ${}(\\{new\_taken}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|t\MG\\{taken}\K\\{new\_taken};{}$\6
\&{if} ${}(\\{new\_taken\_bytes}>\\{old\_taken\_bytes}){}$\5
${}\{{}$\1\6
${}\\{memset}(\|t\MG\\{taken}+\\{old\_taken\_bytes},\39\T{0},\39(\\{new\_taken%
\_bytes}-\\{old\_taken\_bytes}));{}$\6
\4${}\}{}$\2\6
${}\\{memset}(\|t\MG\\{nodes}+\|t\MG\\{capacity},\39\T{0},\39(\\{new%
\_capacity}-\|t\MG\\{capacity})*\&{sizeof}(\&{char}));{}$\6
${}\|t\MG\\{capacity}\K\\{new\_capacity};{}$\6
\\{relink\_trie}(\|t);\6
\&{return} \|t;\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_trie}(\&{struct} \\{trie} ${}{*}\|t)\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\|t\MG\\{nodes});{}$\6
${}\\{free}(\|t\MG\\{links});{}$\6
${}\\{free}(\|t\MG\\{aux});{}$\6
${}\\{free}(\|t\MG\\{taken});{}$\6
\\{free}(\|t);\6
\4${}\}{}$\2\par
\fi

\M{58}relink\_trie.
Connects all empty spaces in the trie.

\Y\B\1\1\&{void} \\{relink\_trie}(\&{struct} \\{trie} ${}{*}\|t)\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{last\_free}${}\K\T{0};{}$\7
\&{for} (\&{size\_t} \\{node}${}\K\T{2};{}$ ${}\\{node}<\|t\MG\\{capacity};{}$
${}\\{node}\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{is\_node\_occupied}(\|t,\39\\{node})){}$\5
${}\{{}$\1\6
${}\\{set\_links}(\|t,\39\\{last\_free},\39\\{node});{}$\6
${}\\{last\_free}\K\\{node};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{set\_links}(\|t,\39\\{last\_free},\39\T{0});{}$\6
\4${}\}{}$\2\par
\fi

\M{59}copy\_node.
Replicates a node between two tries. Return value indicates the success of the
operation.

\Y\B\1\1\&{bool} \\{copy\_node}(\&{struct} \\{trie} ${}{*}\\{from},\39{}$%
\&{size\_t} \\{from\_index}${},\39{}$\&{struct} \\{trie} ${}{*}\\{to},\39{}$%
\&{size\_t} \\{to\_index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{to\_index}\G\\{to}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K((\\{to\_index}/\\{to}\MG\\{capacity})+%
\T{1})*\\{to}\MG\\{capacity};{}$\7
\&{if} ${}(\\{resize\_trie}(\\{to},\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\7
\&{bool} \\{from\_free}${}\K(\\{from}\MG\\{nodes}[\\{from\_index}]\E\T{0});{}$\6
\&{bool} \\{to\_free}${}\K(\\{to}\MG\\{nodes}[\\{to\_index}]\E\T{0});{}$\7
${}\\{to}\MG\\{nodes}[\\{to\_index}]\K\\{from}\MG\\{nodes}[\\{from\_index}];{}$%
\6
${}\\{to}\MG\\{links}[\\{to\_index}]\K\\{from}\MG\\{links}[\\{from\_index}];{}$%
\6
${}\\{to}\MG\\{aux}[\\{to\_index}]\K\\{from}\MG\\{aux}[\\{from\_index}];{}$\6
\&{if} ${}(\R\\{from\_free}\W\\{to\_free}){}$\5
${}\{{}$\1\6
${}\\{to}\MG\\{occupied}\PP;{}$\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{from\_free}\W\R\\{to\_free}){}$\5
${}\{{}$\1\6
${}\\{to}\MG\\{occupied}\MM;{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{get\_base\_used}(\&{struct} \\{trie} ${}{*}\|t,\39{}$\&{size%
\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\|t\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\7
\&{size\_t} \\{byte\_index}${}\K\\{index}/\T{8};{}$\6
\&{size\_t} \\{bit\_index}${}\K\\{index}\MOD\T{8};{}$\7
\&{return} ${}(\|t\MG\\{taken}[\\{byte\_index}]\AND(\T{1}\LL\\{bit\_index}))\I%
\T{0};{}$\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{set\_base\_used}(\&{struct} \\{trie} ${}{*}\|t,\39{}$\&{size%
\_t} \\{index}${},\39{}$\&{bool} \\{used})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\|t\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{resize\_trie}(\|t,\39\\{index}+\T{1})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\7
\&{size\_t} \\{byte\_index}${}\K\\{index}/\T{8};{}$\6
\&{size\_t} \\{bit\_index}${}\K\\{index}\MOD\T{8};{}$\7
\&{if} (\\{used})\5
${}\{{}$\1\6
${}\|t\MG\\{taken}[\\{byte\_index}]\MRL{{\OR}{\K}}(\T{1}\LL\\{bit\_index});{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\|t\MG\\{taken}[\\{byte\_index}]\MRL{\AND{\K}}\CM(\T{1}\LL\\{bit%
\_index});{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{60}set\_links.
Sets the relevant ({\tt link} and {\tt aux}) pointers to link two empty spaces
next to each other.

\Y\B\1\1\&{void} \\{set\_links}(\&{struct} \\{trie} ${}{*}\|t,\39{}$\&{size\_t}
\\{from}${},\39{}$\&{size\_t} \\{to})\2\2\6
${}\{{}$\1\6
${}\|t\MG\\{links}[\\{from}]\K\\{to};{}$\6
${}\|t\MG\\{aux}[\\{to}]\K\\{from};{}$\6
\4${}\}{}$\2\par
\fi

\M{61}is\_node\_occupied.
Returns true if the given trie index is occupied by a node.

\Y\B\1\1\&{bool} \\{is\_node\_occupied}(\&{struct} \\{trie} ${}{*}\|t,\39{}$%
\&{size\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{return} \|t${}\MG\\{nodes}[\\{index}]\I\T{0};{}$\6
\4${}\}{}$\2\par
\fi

\M{62}insert\_pattern.
Attempts to put the given pattern into the trie, possibly creating nodes along
the way. If the insertion is successful,
true is returned and the resulting index is stored in {\tt out\_op\_index}.

\Y\B\1\1\&{bool} \\{insert\_pattern}(\&{struct} \\{trie} ${}{*}\|t,\39{}$%
\&{const} \&{char} ${}{*}\\{pattern},\39{}$\&{size\_t} ${}{*}\\{out\_op%
\_index},\39{}$\&{struct} \\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{length}${}\K\\{strlen}(\\{pattern});{}$\7
\&{return} \\{insert\_substring}${}(\|t,\39\\{pattern},\39\\{length},\39%
\\{length},\39\\{out\_op\_index},\39\\{helper\_trie});{}$\6
\4${}\}{}$\2\par
\fi

\M{63}insert\_substring.
Attempts to put substring $[end-length, end)$ of the given pattern into the
trie, possibly creating nodes along the
way. If the insertion is successful, true is returned and the resulting index
is stored in {\tt out\_op\_index}.

\Y\B\1\1\&{bool} \\{insert\_substring}(\&{struct} \\{trie} ${}{*}\|t,\39{}$%
\&{const} \&{char} ${}{*}\\{pattern},\39{}$\&{size\_t} \\{end}${},\39{}$\&{size%
\_t} \\{length}${},\39{}$\&{size\_t} ${}{*}\\{out\_op\_index},\39{}$\&{struct} %
\\{trie} ${}{*}\\{helper\_trie})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{index}${}\K\\{end}-\\{length};{}$\6
\&{size\_t} \\{base}${}\K\T{1};{}$\6
\&{size\_t} \\{node}${}\K\\{base}+(\\{uint8\_t})\\{pattern}[\\{index}];{}$\6
\&{size\_t} \\{fit};\6
\&{size\_t} \\{node\_prev}${}\K\T{0};{}$\6
\&{bool} \\{new\_pattern}${}\K\\{false};{}$\7
\&{while} ${}(\\{index}<\\{end}\W\\{base}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{node}\K\\{base}+(\\{uint8\_t})\\{pattern}[\\{index}];{}$\6
\&{if} ${}(\\{node}\G\|t\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K((\\{node}/\|t\MG\\{capacity})+\T{1})*\|t\MG%
\\{capacity};{}$\7
\&{if} ${}(\\{resize\_trie}(\|t,\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\|t\MG\\{nodes}[\\{node}]\I\\{pattern}[\\{index}]){}$\5
${}\{{}$\1\6
${}\\{new\_pattern}\K\\{true};{}$\6
\&{if} ${}(\|t\MG\\{nodes}[\\{node}]\E\T{0}){}$\5
${}\{{}$\1\6
${}\|t\MG\\{nodes}[\\{node}]\K\\{pattern}[\\{index}];{}$\6
${}\\{set\_links}(\|t,\39\|t\MG\\{aux}[\\{node}],\39\|t\MG\\{links}[%
\\{node}]);{}$\6
${}\|t\MG\\{aux}[\\{node}]\K\T{0};{}$\6
${}\|t\MG\\{links}[\\{node}]\K\T{0};{}$\6
${}\|t\MG\\{occupied}\PP;{}$\6
\&{if} ${}(\\{node}>\|t\MG\\{node\_max}){}$\5
${}\{{}$\1\6
${}\|t\MG\\{node\_max}\K\\{node};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{if} ${}(\R\\{repack}(\|t,\39\\{helper\_trie},\39{\AND}\\{node\_prev},\39{%
\AND}\\{node},\39\\{pattern}[\\{index}])){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{index}\PP;{}$\6
${}\\{node\_prev}\K\\{node};{}$\6
${}\\{base}\K\|t\MG\\{links}[\\{node}];{}$\6
\4${}\}{}$\2\6
${}\\{helper\_trie}\MG\\{links}[\T{1}]\K\T{0};{}$\6
${}\\{helper\_trie}\MG\\{aux}[\T{1}]\K\T{0};{}$\6
${}\\{helper\_trie}\MG\\{node\_max}\K\T{1};{}$\6
\&{while} ${}(\\{index}<\\{end}){}$\5
${}\{{}$\1\6
${}\\{helper\_trie}\MG\\{nodes}[\T{1}]\K\\{pattern}[\\{index}];{}$\6
\&{if} ${}(\R\\{first\_fit}(\|t,\39\\{helper\_trie},\39{\AND}\\{fit})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\|t\MG\\{links}[\\{node}]\K\\{fit};{}$\6
${}\\{base}\K\\{fit};{}$\6
${}\\{node}\K\\{base}+(\\{uint8\_t})\\{pattern}[\\{index}];{}$\6
${}\\{index}\PP;{}$\6
${}\\{new\_pattern}\K\\{true};{}$\6
\4${}\}{}$\2\6
${}{*}\\{out\_op\_index}\K\\{node};{}$\6
\&{if} (\\{new\_pattern})\5
${}\{{}$\1\6
${}\|t\MG\\{pattern\_count}\PP;{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{64}repack.
Attempts to move a subtree of the trie, so that it fits new node with given
value. If the repacking finishes
successfully, the index for new node is stored in {\tt base} and true is
returned.

\Y\B\1\1\&{bool} \\{repack}(\&{struct} \\{trie} ${}{*}\|t,\39{}$\&{struct} %
\\{trie} ${}{*}\|q,\39{}$\&{size\_t} ${}{*}\\{node},\39{}$\&{size\_t} ${}{*}%
\\{base},\39{}$\&{char} \\{value})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\R\\{unpack}(\|t,\39{*}\\{base}-(\\{uint8\_t})\\{value},\39\|q)){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\|q\MG\\{node\_max}\G\|q\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K((\|q\MG\\{node\_max}/\|q\MG\\{capacity})+%
\T{1})*\|q\MG\\{capacity};{}$\7
\&{if} ${}(\\{resize\_trie}(\|q,\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\|q\MG\\{nodes}[\|q\MG\\{node\_max}]\K\\{value};{}$\6
${}\|q\MG\\{links}[\|q\MG\\{node\_max}]\K\T{0};{}$\6
${}\|q\MG\\{aux}[\|q\MG\\{node\_max}]\K\T{0};{}$\7
\&{size\_t} \\{fit};\7
\&{if} ${}(\R\\{first\_fit}(\|t,\39\|q,\39{\AND}\\{fit})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}{*}\\{base}\K\\{fit};{}$\6
${}\|t\MG\\{links}[{*}\\{node}]\K{*}\\{base};{}$\6
${}{*}\\{base}\MRL{+{\K}}(\\{uint8\_t})\\{value};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{65}unpack.
Moves all nodes with the given base to auxiliary trie.

\Y\B\1\1\&{bool} \\{unpack}(\&{struct} \\{trie} ${}{*}\\{from},\39{}$\&{size%
\_t} \\{base}${},\39{}$\&{struct} \\{trie} ${}{*}\\{to})\2\2{}$\6
${}\{{}$\1\6
${}\\{to}\MG\\{node\_max}\K\T{1};{}$\6
\&{for} (\&{size\_t} \|i${}\K\T{1};{}$ ${}\|i<\T{256};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
\&{size\_t} \\{from\_index}${}\K\\{base}+\|i;{}$\7
\&{if} ${}((\\{uint8\_t})\\{from}\MG\\{nodes}[\\{from\_index}]\E\|i){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{copy\_node}(\\{from},\39\\{from\_index},\39\\{to},\39\\{to}\MG%
\\{node\_max})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{deallocate\_node}(\\{from},\39\\{from\_index});{}$\6
${}\\{to}\MG\\{node\_max}\PP;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{set\_base\_used}(\\{from},\39\\{base},\39\\{false})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{66}first\_fit.
Finds the base of trie {\tt t} that fits the nodes from auxiliary trie {\tt q}
and copies them into their new indices.
Return true if the search and copying finishes successfully.

\Y\B\1\1\&{bool} \\{first\_fit}(\&{struct} \\{trie} ${}{*}\|t,\39{}$\&{struct} %
\\{trie} ${}{*}\|q,\39{}$\&{size\_t} ${}{*}\\{out\_base})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{base};\7
\&{if} ${}(\R\\{find\_base\_for\_first\_fit}(\|t,\39\|q,\39{\AND}\\{base})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \\{q\_index}${}\K\T{1};{}$ ${}\\{q\_index}\Z\|q\MG\\{node%
\_max};{}$ ${}\\{q\_index}\PP){}$\5
${}\{{}$\1\6
\&{size\_t} \\{t\_index}${}\K\\{base}+(\\{uint8\_t})\|q\MG\\{nodes}[\\{q%
\_index}];{}$\7
${}\\{set\_links}(\|t,\39\|t\MG\\{aux}[\\{t\_index}],\39\|t\MG\\{links}[\\{t%
\_index}]);{}$\6
\&{if} ${}(\R\\{copy\_node}(\|q,\39\\{q\_index},\39\|t,\39\\{t\_index})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{if} ${}(\\{t\_index}>\|t\MG\\{node\_max}){}$\5
${}\{{}$\1\6
${}\|t\MG\\{node\_max}\K\\{t\_index};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{set\_base\_used}(\|t,\39\\{base},\39\\{true})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}{*}\\{out\_base}\K\\{base};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{67}find\_base\_for\_first\_fit.
Searches through the trie {\tt t} to find base index that fits the nodes from
the {\tt q} trie. If the search is
successful, resulting index is stored in {\tt out\_base} and true is returned.

\Y\B\1\1\&{bool} \\{find\_base\_for\_first\_fit}(\&{struct} \\{trie} ${}{*}\|t,%
\39{}$\&{struct} \\{trie} ${}{*}\|q,\39{}$\&{size\_t} ${}{*}\\{out\_base})\2%
\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{t\_index};\7
\\{uint8\_t}\\{offset};\6
\&{if} ${}(\|q\MG\\{node\_max}>\T{5}\W\|t\MG\\{capacity}>\|t\MG\\{node\_max}+%
\T{1}){}$\5
${}\{{}$\1\6
${}\\{t\_index}\K\|t\MG\\{node\_max}+\T{1};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{t\_index}\K\|t\MG\\{links}[\T{0}];{}$\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i\Z\|t\MG\\{capacity};{}$ ${}\|i%
\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{t\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{t\_index}\K\|t\MG\\{capacity}+\T{1}{}$;\SHC{ free for sure after resize }%
\7
\&{size\_t} \\{new\_capacity}${}\K\T{2}*\|t\MG\\{capacity};{}$\7
\&{if} ${}(\R\\{resize\_trie}(\|t,\39\\{new\_capacity})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{offset}\K(\\{uint8\_t})\|q\MG\\{nodes}[\T{1}];{}$\6
\&{if} ${}(\\{t\_index}\Z\\{offset}){}$\5
${}\{{}$\1\6
${}\\{t\_index}\K\|t\MG\\{links}[\\{t\_index}];{}$\6
\&{continue};\6
\4${}\}{}$\2\6
${}{*}\\{out\_base}\K\\{t\_index}-\\{offset};{}$\7
\&{size\_t} \\{max\_target\_index}${}\K{*}(\\{out\_base})+\T{255};{}$\7
\&{if} ${}(\\{max\_target\_index}\G\|t\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K((\\{max\_target\_index}/\|t\MG\\{capacity})+%
\T{1})*\|t\MG\\{capacity};{}$\7
\&{if} ${}(\\{resize\_trie}(\|t,\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{get\_base\_used}(\|t,\39{*}\\{out\_base})){}$\5
${}\{{}$\1\6
${}\\{t\_index}\K\|t\MG\\{links}[\\{t\_index}];{}$\6
\&{continue};\6
\4${}\}{}$\2\7
\&{bool} \\{conflict}${}\K\\{false};{}$\7
\&{for} (\&{size\_t} \\{q\_index}${}\K\|q\MG\\{node\_max};{}$ ${}\\{q\_index}\G%
\T{2};{}$ ${}\\{q\_index}\MM){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{is\_node\_occupied}(\|t,\39{*}\\{out\_base}+(\\{uint8\_t})\|q\MG%
\\{nodes}[\\{q\_index}])){}$\5
${}\{{}$\1\6
${}\\{conflict}\K\\{true};{}$\6
\&{break};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{conflict}){}$\5
${}\{{}$\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\6
${}\\{t\_index}\K\|t\MG\\{links}[\\{t\_index}];{}$\6
\4${}\}{}$\2\6
\\{printf}(\.{"Loop\ detected!\\n"});\6
\&{return} \\{false};\6
\4${}\}{}$\2\par
\fi

\M{68}traverse\_trie.
Returns the index of the node corresponding to the given pattern, or 0 if such
node does not exist.

\Y\B\1\1\&{size\_t} \\{traverse\_trie}(\&{struct} \\{trie} ${}{*}\|t,\39{}$%
\&{const} \&{char} ${}{*}\\{pattern})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{index}${}\K\T{1};{}$\6
\&{size\_t} \\{node}${}\K(\\{uint8\_t})\\{pattern}[\T{0}]+\T{1};{}$\6
\&{size\_t} \\{base}${}\K\|t\MG\\{links}[\\{node}];{}$\7
\&{while} ${}(\\{index}<\\{strlen}(\\{pattern})\W\\{base}>\T{0}){}$\5
${}\{{}$\1\6
${}\\{base}\MRL{+{\K}}(\\{uint8\_t})\\{pattern}[\\{index}];{}$\6
\&{if} ${}(\|t\MG\\{nodes}[\\{base}]\I\\{pattern}[\\{index}]){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
${}\\{node}\K\\{base};{}$\6
${}\\{base}\K\|t\MG\\{links}[\\{node}];{}$\6
${}\\{index}\PP;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{index}<\\{strlen}(\\{pattern})){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \\{node};\6
\4${}\}{}$\2\par
\fi

\N{1}{69}Outputs.
The hyphenation information associated to a pattern is stored as a triplet in {%
\tt output} structure:

\item{$\bullet$} {\bf value}: the hyphenation level,
\item{$\bullet$} {\bf position}: the dot position,
\item{$\bullet$} {\bf next\_op\_index}: the pointer to next output of the same
pattern.

For space-saving purposes, several patterns may share a single output. This
means that, for example, for all patterns
having single hyphen of level 1 at position 3 we store only one such output to
which all the patterns point. It is thus
necessary to access the outputs as effectively as possible. \utfpatgen{} uses a
"hash table" with indirect addressing --
simple function $(next\_op\_index + 313*position + 361*value) \% table%
\_capacity) + 1$ computes the index of an output
in the {\tt lookup} array that contains pointers to the real outputs. The whole
structure has these fields:

\item{$\bullet$} {\bf capacity}: the capacity of the array of outputs,
\item{$\bullet$} {\bf count}: the number of outputs currently in the array,
\item{$\bullet$} {\bf data}: the array of outputs,
\item{$\bullet$} {\bf lookup\_cap}: the capacity of the lookup table,
\item{$\bullet$} {\bf lookup\_cnt}: the number of entries currently in the
lookup,
\item{$\bullet$} {\bf lookup}: the array of lookup entries.

The sizes of data and lookup are independent to some extent, so the
corresponding arrays are resized independently.
Since we do not want to fill the lookup fully at any time due to efficiency
reasons, we trigger its resize already at
75 \% capacity. Lookup index 0 does not point to any output and the program
interprets it as "no output present".

\Y\B\1\1\&{struct} \\{outputs} ${}{*}{}$\\{init\_outputs}(\&{size\_t} %
\\{capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{outputs} ${}{*}\\{ops}\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{outputs}));\7
\&{if} ${}(\\{ops}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{ops}\MG\\{capacity}\K\\{capacity};{}$\6
${}\\{ops}\MG\\{count}\K\T{0};{}$\6
${}\\{ops}\MG\\{data}\K\\{calloc}(\\{capacity}+\T{1},\39{}$\&{sizeof}(%
\&{struct} \\{output}));\6
\&{if} ${}(\\{ops}\MG\\{data}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\\{free}(\\{ops});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\T{1};{}$\6
${}\\{ops}\MG\\{lookup\_cap}\K\T{2}*\\{capacity};{}$\6
${}\\{ops}\MG\\{lookup\_cnt}\K\T{0};{}$\6
${}\\{ops}\MG\\{lookup}\K\\{calloc}(\T{2}*\\{capacity}+\T{1},\39\&{sizeof}(%
\&{size\_t}));{}$\6
\&{if} ${}(\\{ops}\MG\\{lookup}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
${}\\{free}(\\{ops}\MG\\{data});{}$\6
\\{free}(\\{ops});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{return} \\{ops};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{outputs} ${}{*}{}$\\{resize\_outputs}(\&{struct} \\{outputs}
${}{*}\\{ops},\39{}$\&{size\_t} \\{capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{output} ${}{*}\\{new\_data}\K\\{realloc}(\\{ops}\MG\\{data},\39(%
\\{capacity}+\T{1})*{}$\&{sizeof}(\&{struct} \\{output}));\7
\&{if} ${}(\\{new\_data}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{ops}\MG\\{data}\K\\{new\_data};{}$\7
\&{size\_t} \\{diff}${}\K\\{capacity}-\\{ops}\MG\\{capacity};{}$\7
${}\\{memset}(\\{ops}\MG\\{data}+\\{ops}\MG\\{capacity}+\T{1},\39\T{0},\39%
\\{diff}*{}$\&{sizeof}(\&{struct} \\{output}));\6
${}\\{ops}\MG\\{capacity}\K\\{capacity};{}$\6
\&{return} \\{ops};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_outputs}(\&{struct} \\{outputs} ${}{*}\\{ops})\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\\{ops}\MG\\{data});{}$\6
${}\\{free}(\\{ops}\MG\\{lookup});{}$\6
\\{free}(\\{ops});\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{resize\_lookup}(\&{struct} \\{outputs} ${}{*}\\{ops},\39{}$%
\&{size\_t} \\{new\_cap}${},\39{}$\&{struct} \\{trie} ${}{*}\|t)\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} ${}{*}\\{new\_lookup}\K\\{calloc}(\\{new\_cap}+\T{1},\39\&{sizeof}(%
\&{size\_t}));{}$\6
\&{size\_t} ${}{*}\\{old\_lookup}\K\\{ops}\MG\\{lookup};{}$\7
\&{if} ${}(\\{new\_lookup}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{ops}\MG\\{lookup}\K\\{new\_lookup};{}$\6
${}\\{ops}\MG\\{lookup\_cap}\K\\{new\_cap};{}$\7
\&{size\_t} \\{old\_hash}${},\39\\{new\_hash},\39\\{op\_index};{}$\6
\&{struct} \\{output} \\{op};\7
\&{for} (\&{size\_t} \\{node}${}\K\T{0};{}$ ${}\\{node}<\|t\MG\\{capacity};{}$
${}\\{node}\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{is\_node\_occupied}(\|t,\39\\{node})\V\|t\MG\\{aux}[\\{node}]\E%
\T{0}){}$\5
${}\{{}$\1\6
\&{continue};\6
\4${}\}{}$\2\6
${}\\{old\_hash}\K\|t\MG\\{aux}[\\{node}];{}$\6
${}\\{op\_index}\K\\{old\_lookup}[\\{old\_hash}];{}$\6
${}\\{op}\K\\{ops}\MG\\{data}[\\{op\_index}];{}$\6
${}\\{new\_hash}\K\\{hash\_trie\_output}(\\{ops},\39\\{op}.\\{value},\39\\{op}.%
\\{position},\39\\{op}.\\{next\_op\_index});{}$\6
${}\\{ops}\MG\\{lookup}[\\{new\_hash}]\K\\{op\_index};{}$\6
${}\|t\MG\\{aux}[\\{node}]\K\\{new\_hash};{}$\6
\4${}\}{}$\2\6
\\{free}(\\{old\_lookup});\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{70}hash\_trie\_output.
Computes the value of the hash function for given output and finds a free index
in the lookup that will store it. If
successful, returns the hash value.

\Y\B\1\1\&{size\_t} \\{hash\_trie\_output}(\&{struct} \\{outputs} ${}{*}%
\\{ops},\39{}$\&{size\_t} \\{value}${},\39{}$\&{size\_t} \\{position}${},\39{}$%
\&{size\_t} \\{next\_op\_index})\2\2\6
${}\{{}$\1\6
\&{size\_t} \\{hash}${}\K((\\{next\_op\_index}+\T{313}*\\{position}+\T{361}*%
\\{value})\MOD\\{ops}\MG\\{lookup\_cap})+\T{1};{}$\6
\&{size\_t} \\{op\_index};\7
\&{while} (\\{true})\5
${}\{{}$\1\6
${}\\{op\_index}\K\\{ops}\MG\\{lookup}[\\{hash}];{}$\6
\&{if} ${}(\\{op\_index}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} \\{hash};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{ops}\MG\\{data}[\\{op\_index}].\\{value}\E\\{value}\W%
\\{ops}\MG\\{data}[\\{op\_index}].\\{position}\E\\{position}\W\\{ops}\MG%
\\{data}[\\{op\_index}].\\{next\_op\_index}\E\\{next\_op\_index}){}$\5
${}\{{}$\1\6
\&{return} \\{hash};\6
\4${}\}{}$\2\6
\&{else} \&{if} ${}(\\{hash}>\T{1}){}$\5
${}\{{}$\1\6
${}\\{hash}\MRL{-{\K}}\T{1};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{hash}\K\\{ops}\MG\\{lookup\_cap};{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{return} \T{0};\6
\4${}\}{}$\2\par
\fi

\N{1}{71}Pattern trie.
The structure that holds already generated set of patterns. It is simply a
aggregation of two other structures:

\item{$\bullet$} {\bf t}: a trie storing the patterns,
\item{$\bullet$} {\bf ops}: outputs array storing the hyphenation information.

Technically speaking, the {\tt aux} pointer of each node that corresponds to a
pattern contains the lookup index of its
output.

\Y\B\1\1\&{struct} \\{pattern\_trie} ${}{*}{}$\\{init\_pattern\_trie}(\&{size%
\_t} \\{trie\_capacity}${},\39{}$\&{size\_t} \\{outputs\_capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{pattern\_trie} ${}{*}\\{pt}\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{pattern\_trie}));\7
\&{if} ${}(\\{pt}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pt}\MG\|t\K\\{init\_trie}(\\{trie\_capacity});{}$\6
\&{if} ${}(\\{pt}\MG\|t\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{pt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{put\_first\_level}(\\{pt}\MG\|t)){}$\5
${}\{{}$\1\6
${}\\{free}(\\{pt}\MG\|t);{}$\6
\\{free}(\\{pt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pt}\MG\\{ops}\K\\{init\_outputs}(\\{outputs\_capacity});{}$\6
\&{if} ${}(\\{pt}\MG\\{ops}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{pt}\MG\|t);{}$\6
\\{free}(\\{pt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{return} \\{pt};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_pattern\_trie}(\&{struct} \\{pattern\_trie} ${}{*}%
\\{pt})\2\2{}$\6
${}\{{}$\1\6
${}\\{destroy\_trie}(\\{pt}\MG\|t);{}$\6
${}\\{destroy\_outputs}(\\{pt}\MG\\{ops});{}$\6
\\{free}(\\{pt});\6
\4${}\}{}$\2\par
\fi

\M{72}new\_trie\_output.
Creates an entry for the given output. If successful, returns true and stores
output's lookup index to {\tt op\_index}.

\Y\B\1\1\&{bool} \\{new\_trie\_output}(\&{struct} \\{pattern\_trie} ${}{*}%
\\{pt},\39{}$\&{size\_t} \\{value}${},\39{}$\&{size\_t} \\{position}${},\39{}$%
\&{size\_t} \\{next\_op\_index}${},\39{}$\&{size\_t} ${}{*}\\{op\_index})$ $\{$
\6
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{count}\G\\{pt}\MG\\{ops}\MG\\{capacity}-%
\T{1}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{resize\_outputs}(\\{pt}\MG\\{ops},\39\\{pt}\MG\\{ops}\MG%
\\{capacity}*\T{2})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{lookup\_cnt}*\T{4}>\\{pt}\MG\\{ops}\MG%
\\{lookup\_cap}*\T{3}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{resize\_lookup}(\\{pt}\MG\\{ops},\39\\{pt}\MG\\{ops}\MG%
\\{lookup\_cap}*\T{2},\39\\{pt}\MG\|t)){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\7
\&{size\_t} \\{hash}${}\K\\{hash\_trie\_output}(\\{pt}\MG\\{ops},\39\\{value},%
\39\\{position},\39\\{next\_op\_index});$ \&{if} ${}(\\{pt}\MG\\{ops}\MG%
\\{lookup}[\\{hash}]\E\T{0})$ $\{$ $\\{pt}\MG\\{ops}\MG\\{count}\PP;$ %
\&{struct} \\{output} \\{new\_op} $\K$ $\{$ $.$ $\\{value}\K\\{value}$ $,$ $.$
$\\{position}\K\\{position}$ $,$ $.$ $\\{next\_op\_index}\K\\{next\_op\_index}$
$\}$  ;\7
\&{size\_t} \\{free\_list\_head}${}\K\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next%
\_op\_index};{}$\7
\&{if} ${}(\\{pt}\MG\\{ops}\MG\\{data}[\\{free\_list\_head}].\\{next\_op%
\_index}\E\T{0}){}$\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\\{pt}\MG\\{ops}\MG%
\\{count}+\T{1};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\T{0}].\\{next\_op\_index}\K\\{pt}\MG\\{ops}\MG%
\\{data}[\\{free\_list\_head}].\\{next\_op\_index};{}$\6
\4${}\}{}$\2\6
${}\\{pt}\MG\\{ops}\MG\\{data}[\\{free\_list\_head}]\K\\{new\_op};{}$\6
${}\\{pt}\MG\\{ops}\MG\\{lookup}[\\{hash}]\K\\{free\_list\_head};{}$\6
${}\\{pt}\MG\\{ops}\MG\\{lookup\_cnt}\PP;$ $\}$ ${*}\\{op\_index}\K\\{hash};{}$%
\6
\&{return} \\{true}; $\}{}$\par
\fi

\M{73}set\_output.
Creates a new output and links it to given node in the trie. Returns true if
the creation and assignment finished
successfully.

\Y\B\1\1\&{bool} \\{set\_output}(\&{struct} \\{pattern\_trie} ${}{*}\\{pt},%
\39{}$\&{size\_t} \\{node}${},\39{}$\&{size\_t} \\{value}${},\39{}$\&{size\_t} %
\\{position})\2\2\6
${}\{{}$\1\6
\&{size\_t} \\{op\_index};\7
\&{if} ${}(\R\\{new\_trie\_output}(\\{pt},\39\\{value},\39\\{position},\39%
\\{pt}\MG\\{ops}\MG\\{lookup}[\\{pt}\MG\|t\MG\\{aux}[\\{node}]],\39{\AND}\\{op%
\_index})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{pt}\MG\|t\MG\\{aux}[\\{node}]\K\\{op\_index};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{74}Pattern counts.
The structure that stores the numbers of supporting and contradicting
occurences for patterns. It comprises 4 fields:

\item{$\bullet$} {\bf capacity}: the maximum index that can be used,
\item{$\bullet$} {\bf size}: the highest index currently in use,
\item{$\bullet$} {\bf good}: array of supporting occurence counts,
\item{$\bullet$} {\bf bad}: array of contradicting occurence counts.

The counts on given index correspond to the same pattern.

\Y\B\1\1\&{struct} \\{pattern\_counts} ${}{*}{}$\\{init\_pattern\_counts}(%
\&{size\_t} \\{capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{pattern\_counts} ${}{*}\\{pc}\K{}$\\{malloc}(\&{sizeof}(%
\&{struct} \\{pattern\_counts}));\7
\&{if} ${}(\\{pc}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pc}\MG\\{good}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{size\_t}));{}$\6
${}\\{pc}\MG\\{bad}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{size\_t}));{}$\6
\&{if} ${}(\\{pc}\MG\\{good}\E\NULL\V\\{pc}\MG\\{bad}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
${}\\{free}(\\{pc}\MG\\{good});{}$\6
${}\\{free}(\\{pc}\MG\\{bad});{}$\6
\\{free}(\\{pc});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pc}\MG\\{capacity}\K\\{capacity};{}$\6
${}\\{pc}\MG\\{size}\K\T{1};{}$\6
\&{return} \\{pc};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{pattern\_counts} ${}{*}{}$\\{resize\_pattern\_counts}(%
\&{struct} \\{pattern\_counts} ${}{*}\\{pc},\39{}$\&{size\_t} \\{new%
\_capacity})\2\2\6
${}\{{}$\1\6
\&{size\_t} ${}{*}\\{new\_good}\K\\{realloc}(\\{pc}\MG\\{good},\39\\{new%
\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_good}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pc}\MG\\{good}\K\\{new\_good};{}$\7
\&{size\_t} ${}{*}\\{new\_bad}\K\\{realloc}(\\{pc}\MG\\{bad},\39\\{new%
\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_bad}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pc}\MG\\{bad}\K\\{new\_bad};{}$\7
\&{size\_t} \\{diff}${}\K\\{new\_capacity}-\\{pc}\MG\\{capacity};{}$\7
${}\\{memset}(\\{pc}\MG\\{good}+\\{pc}\MG\\{capacity},\39\T{0},\39\\{diff}*%
\&{sizeof}(\&{size\_t}));{}$\6
${}\\{memset}(\\{pc}\MG\\{bad}+\\{pc}\MG\\{capacity},\39\T{0},\39\\{diff}*%
\&{sizeof}(\&{size\_t}));{}$\6
${}\\{pc}\MG\\{capacity}\K\\{new\_capacity};{}$\6
\&{return} \\{pc};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_pattern\_counts}(\&{struct} \\{pattern\_counts} ${}{*}%
\\{pc})\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\\{pc}\MG\\{good});{}$\6
${}\\{free}(\\{pc}\MG\\{bad});{}$\6
\\{free}(\\{pc});\6
\4${}\}{}$\2\par
\fi

\N{1}{75}Count trie.
This structure stores candidate patterns and their occurence counts. During
dictionary file processing, every word
contributes to count trie with its substrings as candidate patterns. In
implementation, the count trie is composed from
2 distinct substructures:

\item{$\bullet$} {\bf t}: a trie storing the patterns,
\item{$\bullet$} {\bf cnts}: pattern counts with the numbers of occurences.

Note that the count trie does not need any outputs, since the only pattern
template that the current iteration explores
is defined by {\tt pat\_len} and {\tt pat\_dot} parameters. This allows us to
store pointers to pattern counts in the
{\tt aux} field of occupied trie nodes instead.

\Y\B\1\1\&{struct} \\{count\_trie} ${}{*}{}$\\{init\_count\_trie}(\&{size\_t} %
\\{trie\_capacity}${},\39{}$\&{size\_t} \\{counts\_capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{count\_trie} ${}{*}\\{ct}\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{count\_trie}));\7
\&{if} ${}(\\{ct}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{ct}\MG\|t\K\\{init\_trie}(\\{trie\_capacity});{}$\6
\&{if} ${}(\\{ct}\MG\|t\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{ct});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{put\_first\_level}(\\{ct}\MG\|t)){}$\5
${}\{{}$\1\6
${}\\{free}(\\{ct}\MG\|t);{}$\6
\\{free}(\\{ct});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{ct}\MG\\{cnts}\K\\{init\_pattern\_counts}(\\{counts\_capacity});{}$\6
\&{if} ${}(\\{ct}\MG\\{cnts}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{ct}\MG\|t);{}$\6
\\{free}(\\{ct});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{return} \\{ct};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_count\_trie}(\&{struct} \\{count\_trie} ${}{*}\\{ct})%
\2\2{}$\6
${}\{{}$\1\6
${}\\{destroy\_trie}(\\{ct}\MG\|t);{}$\6
${}\\{destroy\_pattern\_counts}(\\{ct}\MG\\{cnts});{}$\6
\\{free}(\\{ct});\6
\4${}\}{}$\2\par
\fi

\N{1}{76}String buffer.
We use buffer for storing text, usually lines read from input files. The
structure contains following fields:

\item{$\bullet$} {\bf capacity}: the maximum length of the string that the
buffer can hold (including the ending
'0x00'),
\item{$\bullet$} {\bf size}: the current length of string stored in the buffer,
\item{$\bullet$} {\bf data}: the array storing the string,
\item{$\bullet$} {\bf eof}: if the buffer is used to read lines from a file,
this flag indicates whether the end of
file was reached..

\Y\B\1\1\&{struct} \\{string\_buffer} ${}{*}{}$\\{init\_buffer}(\&{size\_t} %
\\{capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{string\_buffer} ${}{*}\\{buf}\K{}$\\{malloc}(\&{sizeof}(%
\&{struct} \\{string\_buffer}));\7
\&{if} ${}(\\{buf}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{capacity}\K\\{capacity};{}$\6
${}\\{buf}\MG\\{size}\K\T{0};{}$\6
${}\\{buf}\MG\\{data}\K\\{malloc}(\\{capacity}*\&{sizeof}(\&{char}));{}$\6
${}\\{buf}\MG\\{eof}\K\\{false};{}$\6
\&{if} ${}(\\{buf}\MG\\{data}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\\{free}(\\{buf});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{data}[\T{0}]\K\.{'\\0'};{}$\6
\&{return} \\{buf};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{string\_buffer} ${}{*}{}$\\{resize\_buffer}(\&{struct} %
\\{string\_buffer} ${}{*}\\{buf},\39{}$\&{size\_t} \\{new\_capacity})\2\2\6
${}\{{}$\1\6
\&{char} ${}{*}\\{new\_ptr}\K\\{realloc}(\\{buf}\MG\\{data},\39\\{new%
\_capacity});{}$\7
\&{if} ${}(\\{new\_ptr}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{data}\K\\{new\_ptr};{}$\6
${}\\{buf}\MG\\{capacity}\K\\{new\_capacity};{}$\6
\&{return} \\{buf};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{reset\_buffer}(\&{struct} \\{string\_buffer} ${}{*}\\{buf})\2%
\2{}$\6
${}\{{}$\1\6
${}\\{buf}\MG\\{eof}\K\\{false};{}$\6
${}\\{buf}\MG\\{size}\K\T{0};{}$\6
${}\\{buf}\MG\\{data}[\T{0}]\K\.{'\\0'};{}$\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_buffer}(\&{struct} \\{string\_buffer} ${}{*}\\{buf})\2%
\2{}$\6
${}\{{}$\1\6
${}\\{free}(\\{buf}\MG\\{data});{}$\6
\\{free}(\\{buf});\6
\4${}\}{}$\2\par
\fi

\M{77}append\_char.
Puts the given character to the current end of buffer. Note that the operation
may overwrite the ending '0x00', and
the caller should whether it is not the case. Returns true upon successful
write.

\Y\B\1\1\&{bool} \\{append\_char}(\&{struct} \\{string\_buffer} ${}{*}\\{buf},%
\39{}$\&{char} \|c)\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{buf}\MG\\{size}+\T{1}\G\\{buf}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{resize\_buffer}(\\{buf},\39\T{2}*\\{buf}\MG\\{capacity})\E%
\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{buf}\MG\\{data}[\\{buf}\MG\\{size}]\K\|c;{}$\6
${}\\{buf}\MG\\{size}\PP;{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{78}append\_string.
Copies the given string to the current end of buffer. Provided string contains
the '0x00' at its end by design, and the
character is copied to the buffer as well. Returns true if the copying finished
successfully.

\Y\B\1\1\&{bool} \\{append\_string}(\&{struct} \\{string\_buffer} ${}{*}%
\\{buf},\39{}$\&{const} \&{char} ${}{*}\\{str})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{len}${}\K\\{strlen}(\\{str})+\T{1};{}$\7
\&{if} ${}(\\{buf}\MG\\{size}+\\{len}\G\\{buf}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{if} ${}(\\{resize\_buffer}(\\{buf},\39\T{2}*(\\{buf}\MG\\{size}+\\{len}))\E%
\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{strcpy}({\AND}\\{buf}\MG\\{data}[\\{buf}\MG\\{size}],\39\\{str});{}$\6
${}\\{buf}\MG\\{size}\MRL{+{\K}}\\{len};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{79}Translate table.
We use this structure to hold the information about character mapping. The
translate file documents the relationships
between "lowercase" letters and their "uppercase" variants, and the translate
table allows easy conversion. Technically
the structure is just a combination of 2 substructures:

\item{$\bullet$} {\bf mapping}: a trie that stores all the characters (both
variants),
\item{$\bullet$} {\bf alphabet}: a string buffer containing all lowercase
characters separated by '0x00'.

In the {\tt aux} field of the trie nodes corresponding to a character we store
index of the beginning of its lowercase
representation in {\tt alphabet}.

\Y\B\1\1\&{struct} \\{translate\_table} ${}{*}{}$\\{init\_tr\_table}(\&{size%
\_t} \\{mapping\_capacity}${},\39{}$\&{size\_t} \\{alphabet\_capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{translate\_table} ${}{*}\\{tt}\K{}$\\{malloc}(\&{sizeof}(%
\&{struct} \\{translate\_table}));\7
\&{if} ${}(\\{tt}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\7
\&{struct} \\{trie} ${}{*}\\{mapping}\K\\{init\_trie}(\\{mapping%
\_capacity});{}$\7
\&{if} ${}(\\{mapping}\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{tt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{put\_first\_level}(\\{mapping})){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{mapping});\6
\\{free}(\\{tt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\7
\&{struct} \\{string\_buffer} ${}{*}\\{alphabet}\K\\{init\_buffer}(\\{alphabet%
\_capacity});{}$\7
\&{if} ${}(\\{alphabet}\E\NULL){}$\5
${}\{{}$\1\6
\\{destroy\_trie}(\\{mapping});\6
\\{free}(\\{tt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{mapping}\K\\{mapping};{}$\6
${}\\{tt}\MG\\{alphabet}\K\\{alphabet};{}$\6
${}\\{tt}\MG\\{index\_to\_alphabet}\K\\{malloc}(\\{alphabet\_capacity}*%
\&{sizeof}(\&{size\_t}));{}$\6
\&{if} ${}(\\{tt}\MG\\{index\_to\_alphabet}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{destroy\_trie}(\\{tt}\MG\\{mapping});{}$\6
${}\\{destroy\_buffer}(\\{tt}\MG\\{alphabet});{}$\6
\\{free}(\\{tt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{tt}\MG\\{letter\_count}\K\T{0};{}$\6
${}\\{tt}\MG\\{letter\_capacity}\K\\{alphabet\_capacity};{}$\6
\&{if} ${}(\R\\{append\_char}(\\{tt}\MG\\{alphabet},\39\.{'\\0'})){}$\5
${}\{{}$\1\6
${}\\{free}(\\{tt}\MG\\{index\_to\_alphabet});{}$\6
${}\\{destroy\_trie}(\\{tt}\MG\\{mapping});{}$\6
${}\\{destroy\_buffer}(\\{tt}\MG\\{alphabet});{}$\6
\\{free}(\\{tt});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{return} \\{tt};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_tr\_table}(\&{struct} \\{translate\_table} ${}{*}%
\\{tt})\2\2{}$\6
${}\{{}$\1\6
${}\\{destroy\_trie}(\\{tt}\MG\\{mapping});{}$\6
${}\\{destroy\_buffer}(\\{tt}\MG\\{alphabet});{}$\6
${}\\{free}(\\{tt}\MG\\{index\_to\_alphabet});{}$\6
\\{free}(\\{tt});\6
\4${}\}{}$\2\par
\fi

\M{80}get\_lower.
Returns the lowercase representation of the given letter, or NULL if the letter
does not exist in the translate table.

\Y\B\1\1\&{char} ${}{*}{}$\\{get\_lower}(\&{struct} \\{translate\_table} ${}{*}%
\\{tt},\39{}$\&{const} \&{char} ${}{*}\\{letter})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{index}${}\K\\{traverse\_trie}(\\{tt}\MG\\{mapping},\39%
\\{letter});{}$\7
\&{if} ${}(\\{index}\E\T{0}\V\\{tt}\MG\\{mapping}\MG\\{aux}[\\{index}]>\\{tt}%
\MG\\{letter\_count}\V\\{tt}\MG\\{mapping}\MG\\{aux}[\\{index}]\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\7
\&{size\_t} \\{alphabet\_offset}${}\K\\{tt}\MG\\{index\_to\_alphabet}[\\{tt}\MG%
\\{mapping}\MG\\{aux}[\\{index}]];{}$\7
\&{if} ${}(\\{alphabet\_offset}\G\\{tt}\MG\\{alphabet}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
\&{return} \\{tt}${}\MG\\{alphabet}\MG\\{data}+\\{alphabet\_offset};{}$\6
\4${}\}{}$\2\par
\fi

\M{81}get\_letter\_index.
Returns the index of the given letter in the translate table, or 0 if the
letter
does not exist.

\Y\B\1\1\&{size\_t} \\{get\_letter\_index}(\&{struct} \\{translate\_table}
${}{*}\\{tt},\39{}$\&{char} ${}{*}\\{letter})\2\2{}$\6
${}\{{}$\1\6
\&{size\_t} \\{index}${}\K\\{traverse\_trie}(\\{tt}\MG\\{mapping},\39%
\\{letter});{}$\7
\&{if} ${}(\\{index}\E\T{0}\V\\{tt}\MG\\{mapping}\MG\\{aux}[\\{index}]>\\{tt}%
\MG\\{letter\_count}\V\\{tt}\MG\\{mapping}\MG\\{aux}[\\{index}]\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \\{tt}${}\MG\\{mapping}\MG\\{aux}[\\{index}];{}$\6
\4${}\}{}$\2\par
\fi

\M{82}convert\_index.
Converts a letter index to a byte sequence and appends it to the word.
While the index is greater than 254, appends 0xFF byte to the word and
subtracts 254 from the index. Finally appends the byte with the same
value as the remainder.

\Y\B\1\1\&{bool} \\{convert\_index}(\&{size\_t} \\{index}${},\39{}$\&{struct} %
\\{word} ${}{*}\\{word})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\&{size\_t} \\{ff\_count}${}\K(\\{index}-\T{1})/\T{254};{}$\6
\&{size\_t} \\{remainder}${}\K((\\{index}-\T{1})\MOD\T{254})+\T{1};{}$\7
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{ff\_count};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{append\_char\_to\_word}(\\{word},\39(\&{char})\,\T{\^ff})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\R\\{append\_char\_to\_word}(\\{word},\39(\&{char})\,%
\\{remainder})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{83}convert\_byte\_sequence.
Converts a byte sequence to a translate table index. This is the reverse
operation
of convert\_index.

\Y\B\1\1\&{size\_t} \\{convert\_byte\_sequence}(\&{char} ${}{*}{*}\\{sequence})%
\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{sequence}\E\NULL\V{*}\\{sequence}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\7
\&{size\_t} \\{ff\_count}${}\K\T{0};{}$\7
\&{while} ${}({*}{*}\\{sequence}\E(\&{char})\,\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{ff\_count}\PP;{}$\6
${}({*}\\{sequence})\PP;{}$\6
\4${}\}{}$\2\7
\&{size\_t} \\{remainder}${}\K(\\{uint8\_t})*{*}\\{sequence};{}$\7
${}({*}\\{sequence})\PP;{}$\6
\&{return} \\{ff\_count}${}*\T{254}+\\{remainder};{}$\6
\4${}\}{}$\2\par
\fi

\N{1}{84}Params.
This structure encompasses all the parameters that influence the computation of
\utfpatgen. We can sort them by their
scope of into {\bf global} (usually defined once and used throughout the whole
run), {\bf level-specific} (defined
specificallly for each hyphenation level), and {\bf pass-specific} (defined
specifically for each iteration within
a level):

\item{$\bullet$} {\bf left\_hyphen\_min} (global): the length of prefix of a
word that should not be hyphenated,
\item{$\bullet$} {\bf right\_hyphen\_min} (global): the length of suffix of a
word that should not be hyphenated,
\item{$\bullet$} {\bf bad\_hyphen} (global): the symbol representing
incorrectly placed hyphen in the dictionary,
\item{$\bullet$} {\bf missed\_hyphen} (global): the symbol representing missing
hyphen in the dictionary,
\item{$\bullet$} {\bf good\_hyphen} (global): the symbol representing correctly
placed hyphen in the dictionary,
\item{$\bullet$} {\bf hyph\_start} (global): the lowest hyphenation level to be
generated,
\item{$\bullet$} {\bf hyph\_finish} (global): the highest hyphenation level to
be generated,
\item{$\bullet$} {\bf word\_weight} (global): the default weight of a hyphen in
a word from dictionary,
\item{$\bullet$} {\bf dictionary\_file} (global): a pointer to opened file with
dictionary entries,
\item{$\bullet$} {\bf pattern\_file} (global): a pointer to opened file with
initial set of patterns,
\item{$\bullet$} {\bf output\_file} (global): a pointer to opened file where
the final set of patterns will be
written,
\item{$\bullet$} {\bf translate\_file} (global): a pointer to opened file with
character mapping and
language-specific parameters,
\item{$\bullet$} {\bf hyph\_level} (level-specific): the current hyphenation
level,
\item{$\bullet$} {\bf pat\_start} (level-specific): the shortest pattern
length,
\item{$\bullet$} {\bf pat\_finish} (level-specific): the longest pattern
length,
\item{$\bullet$} {\bf good\_wt} (level-specific): the supporting occurence
weight,
\item{$\bullet$} {\bf bad\_wt} (level-specific): the contradicting occurence
weight,
\item{$\bullet$} {\bf thresh} (level-specific): the pattern acceptance
threshold,
\item{$\bullet$} {\bf good\_dot} (level-specific): the hyphen type considered
as correct,
\item{$\bullet$} {\bf bad\_dot} (level-specific): the hyphen type considered as
incorrect,
\item{$\bullet$} {\bf pat\_len} (pass-specific): the current pattern length,
\item{$\bullet$} {\bf pat\_dot} (pass-specific): the current dot position.

\Y\B\1\1\&{struct} \\{params} ${}{*}\\{init\_params}(\,)\2\2{}$\6
${}\{{}$\1\6
\&{struct} \\{params} ${}{*}\|p\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{params}));\7
\&{if} ${}(\|p\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fputs}(\.{"Allocation\ error\\n"},\39\\{stderr});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|p\MG\\{left\_hyphen\_min}\K\T{2};{}$\6
${}\|p\MG\\{right\_hyphen\_min}\K\T{3};{}$\6
${}\|p\MG\\{bad\_hyphen}\K\.{'.'};{}$\6
${}\|p\MG\\{missed\_hyphen}\K\.{'-'};{}$\6
${}\|p\MG\\{good\_hyphen}\K\.{'*'};{}$\6
${}\|p\MG\\{word\_weight}\K\T{1};{}$\6
${}\|p\MG\\{dictionary\_file}\K\NULL;{}$\6
${}\|p\MG\\{pattern\_file}\K\NULL;{}$\6
${}\|p\MG\\{output\_file}\K\NULL;{}$\6
${}\|p\MG\\{translate\_file}\K\NULL;{}$\6
\&{return} \|p;\6
\4${}\}{}$\2\7
\1\1\&{void} \\{reset\_params}(\&{struct} \\{params} ${}{*}\|p)\2\2{}$\6
${}\{{}$\1\6
${}\|p\MG\\{left\_hyphen\_min}\K\T{2};{}$\6
${}\|p\MG\\{right\_hyphen\_min}\K\T{3};{}$\6
${}\|p\MG\\{bad\_hyphen}\K\.{'.'};{}$\6
${}\|p\MG\\{missed\_hyphen}\K\.{'-'};{}$\6
${}\|p\MG\\{good\_hyphen}\K\.{'*'};{}$\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_params}(\&{struct} \\{params} ${}{*}\|p)\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\|p\MG\\{dictionary\_file}\I\NULL){}$\5
${}\{{}$\1\6
${}\\{fclose}(\|p\MG\\{dictionary\_file});{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|p\MG\\{pattern\_file}\I\NULL){}$\5
${}\{{}$\1\6
${}\\{fclose}(\|p\MG\\{pattern\_file});{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|p\MG\\{output\_file}\I\NULL){}$\5
${}\{{}$\1\6
${}\\{fclose}(\|p\MG\\{output\_file});{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\|p\MG\\{translate\_file}\I\NULL){}$\5
${}\{{}$\1\6
${}\\{fclose}(\|p\MG\\{translate\_file});{}$\6
\4${}\}{}$\2\6
\\{free}(\|p);\6
\4${}\}{}$\2\par
\fi

\N{1}{85}Pass stats.
This structure groups together variables with mostly statistical meaning. We
use them to map the progress and quality
of the pattern generation, and present it on standard output. Following fields
are monitored:

\item{$\bullet$} {\bf good\_pat\_cnt}: the number of new good patterns from an
iteration,
\item{$\bullet$} {\bf bad\_pat\_cnt}: the number of new bad patterns from an
iteration,
\item{$\bullet$} {\bf good\_cnt}: the number of hyphens from the dictionary
correctly identified with current set
of patterns,
\item{$\bullet$} {\bf bad\_cnt}: the number of hyphens from the dictionary
incorrectly identified with current set
of patterns,
\item{$\bullet$} {\bf miss\_cnt}: the number of hyphens from the dictionary
missed with current set of patterns,
\item{$\bullet$} {\bf level\_pattern\_cnt}: the number of patterns generated on
current hyphenation level,
\item{$\bullet$} {\bf max\_level}: the highest hyphenation level found in the
initial patterns,
\item{$\bullet$} {\bf more\_to\_come}: a flag indicating that undecided
patterns were found on current hyphenation
level.

\fi

\N{1}{86}Stack.
We decided to implement a stack-like structure to break the recursive functions
from \patgen. In most cases the stack
holds data related to previous iterations of a method, for instance, the trie
bases on the path to current node. There
are only 3 fields:

\item{$\bullet$} {\bf capacity}: the highest index that can be accessed,
\item{$\bullet$} {\bf top}: the index where the last value was inserted,
\item{$\bullet$} {\bf data}: the array of values currently on the stack.

\Y\B\1\1\&{struct} \\{stack} ${}{*}{}$\\{init\_stack}(\&{size\_t} \\{capacity})%
\2\2\6
${}\{{}$\1\6
\&{struct} \\{stack} ${}{*}\|s\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{stack}));\7
\&{if} ${}(\|s\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|s\MG\\{data}\K\\{malloc}(\\{capacity}*\&{sizeof}(\&{size\_t}));{}$\6
\&{if} ${}(\|s\MG\\{data}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
${}\\{free}(\|s\MG\\{data});{}$\6
\\{free}(\|s);\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|s\MG\\{capacity}\K\\{capacity};{}$\6
${}\|s\MG\\{top}\K\T{0};{}$\6
\&{return} \|s;\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{stack} ${}{*}{}$\\{resize\_stack}(\&{struct} \\{stack} ${}{*}%
\|s,\39{}$\&{size\_t} \\{new\_capacity})\2\2\6
${}\{{}$\1\6
\&{size\_t} ${}{*}\\{new\_stack}\K\\{realloc}(\|s\MG\\{data},\39\\{new%
\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_stack}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\\{destroy\_stack}(\|s);\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\|s\MG\\{data}\K\\{new\_stack};{}$\6
${}\|s\MG\\{capacity}\K\\{new\_capacity};{}$\6
\&{return} \|s;\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_stack}(\&{struct} \\{stack} ${}{*}\|s)\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\|s\MG\\{data});{}$\6
\\{free}(\|s);\6
\4${}\}{}$\2\par
\fi

\M{87}put\_on\_stack.
Appends the given value to the top of the stack. Returns true if the insertion
finishes successfully.

\Y\B\1\1\&{bool} \\{put\_on\_stack}(\&{struct} \\{stack} ${}{*}\|s,\39{}$%
\&{size\_t} \\{value})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\|s\MG\\{top}\G\|s\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{size\_t} \\{new\_capacity}${}\K\T{2}*(\|s\MG\\{top});{}$\7
\&{if} ${}(\\{resize\_stack}(\|s,\39\\{new\_capacity})\E\NULL){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\|s\MG\\{data}[\|s\MG\\{top}]\K\\{value};{}$\6
${}\|s\MG\\{top}\PP;{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{88}get\_top\_value.
Returns the value currently at {\tt top} index. Does not change anything on the
stack.

\Y\B\1\1\&{size\_t} \\{get\_top\_value}(\&{struct} \\{stack} ${}{*}\|s)\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\|s\MG\\{top}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \|s${}\MG\\{data}[\|s\MG\\{top}-\T{1}];{}$\6
\4${}\}{}$\2\par
\fi

\M{89}set\_top\_value.
Changes the value at {\tt top} index to the given one.

\Y\B\1\1\&{void} \\{set\_top\_value}(\&{struct} \\{stack} ${}{*}\|s,\39{}$%
\&{size\_t} \\{value})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\|s\MG\\{top}\E\T{0}){}$\5
${}\{{}$\1\6
\&{return};\6
\4${}\}{}$\2\6
${}\|s\MG\\{data}[\|s\MG\\{top}-\T{1}]\K\\{value};{}$\6
\4${}\}{}$\2\par
\fi

\N{1}{90}Word.
The structure that holds information about a word and its hyphens. It consists
of 7 fields:

\item{$\bullet$} {\bf capacity}: the maximum length of the word in bytes,
\item{$\bullet$} {\bf size}: current length of the word in bytes,
\item{$\bullet$} {\bf length}: current length of the word in characters,
\item{$\bullet$} {\bf lowercase}: the word translated to lowercase characters
according to the translate table, and
surrounded by {\tt EDGE\_OF\_WORD} symbols,
\item{$\bullet$} {\bf true\_hyphens}: array of hyphen types and their weights
retrieved from the dictionary,
\item{$\bullet$} {\bf found\_hyphens}: array of hyphenation level identified by
current set of patterns,
\item{$\bullet$} {\bf no\_more}: array of flags marking the dot positions that
can be skipped during processing.

Array index $n$ corresponds to the dot position between characters $n$ and
$n+1$. The lowest two bits of a
{\tt true\_hyphens} value describe the hyphen type, the rest represent its
weight (multiplied by 4).

Note that the {\tt lowercase} array is not strictly a string as we do not
require the closing '0x00'.

\Y\B\1\1\&{struct} \\{word} ${}{*}{}$\\{init\_word}(\&{size\_t} \\{capacity})\2%
\2\6
${}\{{}$\1\6
\&{struct} \\{word} ${}{*}\\{word}\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{word}));\7
\&{if} ${}(\\{word}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{translated}\K\\{calloc}(\\{capacity},\39\&{sizeof}(%
\&{char}));{}$\6
\&{if} ${}(\\{word}\MG\\{translated}\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{word});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{true\_hyphens}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{size%
\_t}));{}$\6
\&{if} ${}(\\{word}\MG\\{true\_hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{word}\MG\\{translated});{}$\6
\\{free}(\\{word});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{found\_hyphens}\K\\{calloc}(\\{capacity},\39{}$\&{sizeof} (%
\\{uint8\_t}));\6
\&{if} ${}(\\{word}\MG\\{found\_hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{word}\MG\\{translated});{}$\6
${}\\{free}(\\{word}\MG\\{true\_hyphens});{}$\6
\\{free}(\\{word});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{no\_more}\K\\{calloc}(\\{capacity},\39\&{sizeof}(%
\&{bool}));{}$\6
\&{if} ${}(\\{word}\MG\\{no\_more}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{word}\MG\\{translated});{}$\6
${}\\{free}(\\{word}\MG\\{true\_hyphens});{}$\6
${}\\{free}(\\{word}\MG\\{found\_hyphens});{}$\6
\\{free}(\\{word});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{size}\K\T{0};{}$\6
${}\\{word}\MG\\{length}\K\T{0};{}$\6
${}\\{word}\MG\\{capacity}\K\\{capacity};{}$\6
\&{return} \\{word};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{word} ${}{*}{}$\\{resize\_word}(\&{struct} \\{word} ${}{*}%
\\{word},\39{}$\&{size\_t} \\{new\_capacity})\2\2\6
${}\{{}$\1\6
\&{char} ${}{*}\\{new\_translated}\K\\{realloc}(\\{word}\MG\\{translated},\39%
\\{new\_capacity}*\&{sizeof}(\&{char}));{}$\7
\&{if} ${}(\\{new\_translated}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{translated}\K\\{new\_translated};{}$\7
\&{size\_t} ${}{*}\\{new\_true\_hyphens}\K\\{realloc}(\\{word}\MG\\{true%
\_hyphens},\39\\{new\_capacity}*\&{sizeof}(\&{size\_t}));{}$\7
\&{if} ${}(\\{new\_true\_hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{true\_hyphens}\K\\{new\_true\_hyphens};{}$\6
${}\\{uint8\_t}*\\{new\_found\_hyphens}\K\\{realloc}(\\{word}\MG\\{found%
\_hyphens},\39\\{new\_capacity}*{}$\&{sizeof} (\\{uint8\_t}));\6
\&{if} ${}(\\{new\_found\_hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{found\_hyphens}\K\\{new\_found\_hyphens};{}$\7
\&{bool} ${}{*}\\{new\_no\_more}\K\\{realloc}(\\{word}\MG\\{no\_more},\39\\{new%
\_capacity}*\&{sizeof}(\&{bool}));{}$\7
\&{if} ${}(\\{new\_no\_more}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{no\_more}\K\\{new\_no\_more};{}$\7
\&{size\_t} \\{diff}${}\K\\{new\_capacity}-\\{word}\MG\\{capacity};{}$\7
${}\\{memset}(\\{word}\MG\\{translated}+\\{word}\MG\\{capacity},\39\.{'\\0'},%
\39\\{diff}*\&{sizeof}(\&{char}));{}$\6
${}\\{memset}(\\{word}\MG\\{true\_hyphens}+\\{word}\MG\\{capacity},\39\T{0},\39%
\\{diff}*\&{sizeof}(\&{size\_t}));{}$\6
${}\\{memset}(\\{word}\MG\\{found\_hyphens}+\\{word}\MG\\{capacity},\39\T{0},%
\39\\{diff}*{}$\&{sizeof} (\\{uint8\_t}));\6
${}\\{memset}(\\{word}\MG\\{no\_more}+\\{word}\MG\\{capacity},\39\\{false},\39%
\\{diff}*\&{sizeof}(\&{bool}));{}$\6
${}\\{word}\MG\\{capacity}\K\\{new\_capacity};{}$\6
\&{return} \\{word};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{reset\_word}(\&{struct} \\{word} ${}{*}\\{word})\2\2{}$\6
${}\{{}$\1\6
${}\\{word}\MG\\{length}\K\T{0};{}$\6
${}\\{word}\MG\\{size}\K\T{0};{}$\6
${}\\{memset}(\\{word}\MG\\{translated},\39\T{0},\39\\{word}\MG\\{capacity}*%
\&{sizeof}(\&{char}));{}$\6
${}\\{memset}(\\{word}\MG\\{true\_hyphens},\39\T{0},\39\\{word}\MG\\{capacity}*%
\&{sizeof}(\&{size\_t}));{}$\6
${}\\{memset}(\\{word}\MG\\{found\_hyphens},\39\T{0},\39\\{word}\MG%
\\{capacity}*{}$\&{sizeof} (\\{uint8\_t}));\6
${}\\{memset}(\\{word}\MG\\{no\_more},\39\\{false},\39\\{word}\MG\\{capacity}*%
\&{sizeof}(\&{bool}));{}$\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_word}(\&{struct} \\{word} ${}{*}\\{word})\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\\{word}\MG\\{translated});{}$\6
${}\\{free}(\\{word}\MG\\{true\_hyphens});{}$\6
${}\\{free}(\\{word}\MG\\{found\_hyphens});{}$\6
${}\\{free}(\\{word}\MG\\{no\_more});{}$\6
\\{free}(\\{word});\6
\4${}\}{}$\2\7
\1\1\&{size\_t} \\{get\_true\_hyphen}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{size\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \\{word}${}\MG\\{true\_hyphens}[\\{index}];{}$\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{set\_true\_hyphen}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{size\_t} \\{index}${},\39{}$\&{size\_t} \\{value})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{true\_hyphens}[\\{index}]\K\\{value};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\1\1\\{uint8\_t}\\{get\_found\_hyphen}(\&{struct} \\{word} ${}{*}\\{word},%
\39{}$\&{size\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \\{word}${}\MG\\{found\_hyphens}[\\{index}];{}$\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{set\_found\_hyphen}(\&{struct} \\{word} ${}{*}\\{word},\39{}$%
\&{size\_t} \\{index}${},\39\\{uint8\_t}\\{value})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{found\_hyphens}[\\{index}]\K\\{value};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{get\_no\_more}(\&{struct} \\{word} ${}{*}\\{word},\39{}$\&{size%
\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\&{return} \\{word}${}\MG\\{no\_more}[\\{index}];{}$\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{set\_no\_more}(\&{struct} \\{word} ${}{*}\\{word},\39{}$\&{size%
\_t} \\{index}${},\39{}$\&{bool} \\{value})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{word}\MG\\{size}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{no\_more}[\\{index}]\K\\{value};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{91}append\_char\_to\_word.
Puts the given byte to the end of word. Returns true if the insertion was
successful.

\Y\B\1\1\&{bool} \\{append\_char\_to\_word}(\&{struct} \\{word} ${}{*}\\{word},%
\39{}$\&{char} \|c)\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{word}\MG\\{size}\G\\{word}\MG\\{capacity}-\T{1}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{resize\_word}(\\{word},\39\T{2}*\\{word}\MG\\{capacity})){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{word}\MG\\{translated}[\\{word}\MG\\{size}]\K\|c;{}$\6
${}\\{word}\MG\\{size}\PP;{}$\6
\&{if} ${}((\\{uint8\_t})\|c\I\T{\^ff}){}$\5
${}\{{}$\1\6
${}\\{word}\MG\\{length}\PP;{}$\6
\4${}\}{}$\2\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{1}{92}Pattern.
This structure's purpose is to hold information about a pattern parsed from the
pattern file. It has similar fields as
{\tt word}, but is stripped of the arrays it does not need:

\item{$\bullet$} {\bf capacity}: the maximum length of the pattern in bytes,
\item{$\bullet$} {\bf size}: current length of the pattern in bytes,
\item{$\bullet$} {\bf length}: current length of the pattern in characters,
\item{$\bullet$} {\bf text}: the pattern translated to lowercase characters
according to the translate table,
\item{$\bullet$} {\bf hyphens}: array of hyphenation levels of the pattern.

\Y\B\1\1\&{struct} \\{pattern} ${}{*}{}$\\{init\_pattern}(\&{size\_t} %
\\{capacity})\2\2\6
${}\{{}$\1\6
\&{struct} \\{pattern} ${}{*}\\{pat}\K{}$\\{malloc}(\&{sizeof}(\&{struct} %
\\{pattern}));\7
\&{if} ${}(\\{pat}\E\NULL){}$\5
${}\{{}$\1\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{text}\K\\{calloc}(\\{capacity},\39\&{sizeof}(\&{char}));{}$\6
\&{if} ${}(\\{pat}\MG\\{text}\E\NULL){}$\5
${}\{{}$\1\6
\\{free}(\\{pat});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{hyphens}\K\\{calloc}(\\{capacity},\39{}$\&{sizeof} (\\{uint8%
\_t}));\6
\&{if} ${}(\\{pat}\MG\\{hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{free}(\\{pat}\MG\\{text});{}$\6
\\{free}(\\{pat});\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{length}\K\T{0};{}$\6
${}\\{pat}\MG\\{size}\K\T{0};{}$\6
${}\\{pat}\MG\\{capacity}\K\\{capacity};{}$\6
\&{return} \\{pat};\6
\4${}\}{}$\2\7
\1\1\&{struct} \\{pattern} ${}{*}{}$\\{resize\_pattern}(\&{struct} \\{pattern}
${}{*}\\{pat},\39{}$\&{size\_t} \\{new\_capacity})\2\2\6
${}\{{}$\1\6
\&{char} ${}{*}\\{new\_text}\K\\{realloc}(\\{pat}\MG\\{text},\39\\{new%
\_capacity}*\&{sizeof}(\&{char}));{}$\7
${}\\{uint8\_t}*\\{new\_hyphens}\K\\{realloc}(\\{pat}\MG\\{hyphens},\39\\{new%
\_capacity}*{}$\&{sizeof} (\\{uint8\_t}));\6
\&{if} ${}(\\{new\_text}\E\NULL\V\\{new\_hyphens}\E\NULL){}$\5
${}\{{}$\1\6
${}\\{fprintf}(\\{stderr},\39\.{"Allocation\ error\\n"});{}$\6
\&{return} ${}\NULL;{}$\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{text}\K\\{new\_text};{}$\6
${}\\{pat}\MG\\{hyphens}\K\\{new\_hyphens};{}$\6
${}\\{memset}(\\{pat}\MG\\{text}+\\{pat}\MG\\{capacity},\39\.{'\\0'},\39(\\{new%
\_capacity}-\\{pat}\MG\\{capacity})*\&{sizeof}(\&{char}));{}$\6
${}\\{memset}(\\{pat}\MG\\{hyphens}+\\{pat}\MG\\{capacity},\39\T{0},\39(\\{new%
\_capacity}-\\{pat}\MG\\{capacity})*{}$\&{sizeof} (\\{uint8\_t}));\6
${}\\{pat}\MG\\{capacity}\K\\{new\_capacity};{}$\6
\&{return} \\{pat};\6
\4${}\}{}$\2\7
\1\1\&{void} \\{reset\_pattern}(\&{struct} \\{pattern} ${}{*}\\{pat})\2\2{}$\6
${}\{{}$\1\6
${}\\{pat}\MG\\{length}\K\T{0};{}$\6
${}\\{pat}\MG\\{size}\K\T{0};{}$\6
${}\\{memset}(\\{pat}\MG\\{text},\39\.{'\\0'},\39\\{pat}\MG\\{capacity}*%
\&{sizeof}(\&{char}));{}$\6
${}\\{memset}(\\{pat}\MG\\{hyphens},\39\T{0},\39\\{pat}\MG\\{capacity}*{}$%
\&{sizeof} (\\{uint8\_t}));\6
\4${}\}{}$\2\7
\1\1\&{void} \\{destroy\_pattern}(\&{struct} \\{pattern} ${}{*}\\{pat})\2\2{}$\6
${}\{{}$\1\6
${}\\{free}(\\{pat}\MG\\{text});{}$\6
${}\\{free}(\\{pat}\MG\\{hyphens});{}$\6
\\{free}(\\{pat});\6
\4${}\}{}$\2\7
\1\1\\{uint8\_t}\\{get\_hyphen}(\&{struct} \\{pattern} ${}{*}\\{pat},\39{}$%
\&{size\_t} \\{index})\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{pat}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{return} \T{0};\6
\4${}\}{}$\2\6
\&{return} \\{pat}${}\MG\\{hyphens}[\\{index}];{}$\6
\4${}\}{}$\2\7
\1\1\&{bool} \\{set\_hyphen}(\&{struct} \\{pattern} ${}{*}\\{pat},\39{}$\&{size%
\_t} \\{index}${},\39\\{uint8\_t}\\{value})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\G\\{pat}\MG\\{capacity}){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{hyphens}[\\{index}]\K\\{value};{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\M{93}convert\_index\_to\_pattern.
Converts a letter index to a byte sequence and appends it to the pattern.
While the index is greater than 254, appends 0xFF byte to the pattern and
subtracts 254 from the index. Finally appends the byte with the same
value as the remainder.

\Y\B\1\1\&{bool} \\{convert\_index\_to\_pattern}(\&{size\_t} \\{index}${},%
\39{}$\&{struct} \\{pattern} ${}{*}\\{pat})\2\2{}$\6
${}\{{}$\1\6
\&{if} ${}(\\{index}\E\T{0}){}$\1\5
\&{return} \\{true};\2\7
\&{size\_t} \\{ff\_count}${}\K(\\{index}-\T{1})/\T{254};{}$\6
\&{size\_t} \\{remainder}${}\K((\\{index}-\T{1})\MOD\T{254})+\T{1};{}$\6
\&{size\_t} \\{total\_bytes}${}\K\\{ff\_count}+\T{1};{}$\7
\&{if} ${}(\\{pat}\MG\\{size}\G\\{pat}\MG\\{capacity}-\\{total\_bytes}){}$\5
${}\{{}$\1\6
\&{if} ${}(\R\\{resize\_pattern}(\\{pat},\39\T{2}*(\\{pat}\MG\\{capacity}+%
\\{total\_bytes}))){}$\5
${}\{{}$\1\6
\&{return} \\{false};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{for} (\&{size\_t} \|i${}\K\T{0};{}$ ${}\|i<\\{ff\_count};{}$ ${}\|i\PP){}$\5
${}\{{}$\1\6
${}\\{pat}\MG\\{text}[\\{pat}\MG\\{size}]\K(\&{char})\,\T{\^ff};{}$\6
${}\\{pat}\MG\\{size}\PP;{}$\6
\4${}\}{}$\2\6
${}\\{pat}\MG\\{text}[\\{pat}\MG\\{size}]\K(\&{char})\,\\{remainder};{}$\6
${}\\{pat}\MG\\{size}\PP;{}$\6
\&{return} \\{true};\6
\4${}\}{}$\2\par
\fi

\N{0}{94}Index.
List of used identifiers
\fi

\inx
\fin
\con
